---
title: "Clean peaktable"
author: "`r params$author`"
date: "`r base::Sys.Date()`"

knit: ( function(inputFile, encoding) { 
          rmarkdown::render( inputFile, 
                             encoding = encoding, 
                             output_file = paste0( substr(inputFile,1,nchar(inputFile)-4), '_', Sys.Date(),'.html')) 
          })

output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '4'
    df_print: paged
---

<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE}

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

```

```{r set path, include = FALSE}

path.output <-
  paste0(params$path.result, "03_Clean_peaktable/")

if (!dir.exists(path.output)) dir.create(path.output)

if (is.null(params$mz_col_nr)){

  params$mz_col_nr <- 2
  params$rt_col_nr <- 3
  params$identity_col_nr <- 4
  
}

if (is.null(params$path.peaktable.pos)){
  
  params$path.peaktable.pos <- paste0(params$path.result, "02_Preprocessing/peaktable_pos.csv")
  
} 

if (is.null(params$path.peaktable.neg)){
  
  params$path.peaktable.neg <- paste0(params$path.result, "02_Preprocessing/peaktable_neg.csv")
  
} 

```

# 1 Positive mode
## 1.1 Data import
```{r data import pos, eval = params$run.pos}

if (is.data.frame(params$path.peaktable.pos)){
  
  peaktable <- params$path.peaktable.pos
  
} else if ( grepl(".xlsx", params$path.peaktable.pos) ){
  
  peaktable <-
    readxl::read_excel( params$path.peaktable.pos )
  
} else {
  
  peaktable <-
    read.csv( params$path.peaktable.pos, 
              sep = params$peaktable.sep, 
              check.names = FALSE)
  
}

print("POS peaktable was imported from the following file:", quote = FALSE)
print(params$path.peaktable.pos, quote = FALSE)

peaks.names <-
  paste0(params$sample.info.pos$Sample.name, params$peaktable.sample.pattern)
 
peaks.sample <-
  peaktable[, peaks.names] %>%
  mutate_if(is.character, as.numeric)

peaks.all <-
  cbind(peaktable[, c(params$mz_col_nr, params$rt_col_nr, params$identity_col_nr)], peaks.sample)

f.info <- 
  peaktable[, c(params$identity_col_nr, params$mz_col_nr, params$rt_col_nr, params$others), drop = FALSE] %>%
  dplyr::rename( Identity = 1,
                 mz = 2,
                 rt = 3) %>%
  mutate(  Identity = ifelse( grepl(": 0.", Identity, fixed = TRUE),
                               substr(Identity, 1, regexpr("\\:[^\\:]*$", Identity)-1),
                               Identity),
           Identity = ifelse( is.na(Identity),
                             paste0("MZ_", round(mz, 3), "_RT_", round(rt, 3)),
                             Identity ),
          Feature.id = row_number()) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  arrange(Feature.id) %>%
  select( -c(n, Feature.id, mz, rt) )

peaks.temp <-
  peaks.all %>%
  dplyr::rename( mz = 1,
                 rt = 2,
                 Identity = 3) %>%
  mutate(  Identity = ifelse( grepl(": 0.", Identity, fixed = TRUE),
                               substr(Identity, 1, regexpr("\\:[^\\:]*$", Identity)-1),
                               Identity),
           Identity = ifelse( is.na(Identity),
                             paste0("MZ_", round(mz, 3), "_RT_", round(rt, 3)),
                             Identity ),
          Feature.id = row_number()) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  arrange(Feature.id) %>%
  select( -c(n, Feature.id) ) %>%
  mutate( Identity = ifelse(nchar(Identity) > 1000, substr(Identity, 1, 1000), Identity)) 

peaks.is <-
  peaks.temp %>%
  filter(grepl("^[0-9][0-9]\\.", substr(Identity, 1, 3)))

peaks <-
  peaks.temp %>%
  filter(!grepl("^[0-9][0-9]\\.", substr(Identity, 1, 3)))

datalist.pos <- list(peaks.raw = peaks,
                     peaks.is.raw = peaks.is)

```

## 1.2 Feature filtering
```{r filtering, eval = params$run.pos}
#### 2 Feature filtering
##### 2.1 Feature inspection
##### 2.2 Filtering
# Peaks
mzrt.filtered <-
  filter_peaks( peaktable = peaks,
                sample.type = params$sample.info.pos$Sample.type,
                rsd.filter = params$rsd.filter,
                rt.range = params$rt.range,
                mean.filter = params$mean.filter)

peaks.filtered <- 
  mzrt.filtered$peaktable

features <- 
  cbind( peaks.filtered[, c("mz", "rt", "Identity")], mzrt.filtered$summary[, -1])

data <- 
  t(peaks.filtered[, !colnames(peaks.filtered) %in% c("mz", "rt", "Identity")]) %>%
  as.data.frame()

colnames(data) <- features$Identity

# IS
if (nrow(peaks.is > 0)){
  mzrt.is.filtered <-
    filter_peaks( peaktable = peaks.is,
                  sample.type = params$sample.info.pos$Sample.type)
  
  peaks.is.filtered <- 
    mzrt.is.filtered$peaktable
  
  
  features.is <- 
    cbind( peaks.is.filtered[, c("mz", "rt", "Identity")], mzrt.is.filtered$summary[, -1])
  
  data.is <- 
    t(peaks.is.filtered[, !colnames(peaks.is.filtered) %in% c("mz", "rt", "Identity")]) %>%
    as.data.frame()
  
  colnames(data.is) <- features.is$Identity
}
```

## 1.3 Filter based on missings
### Feature
```{r filter, eval = params$run.pos}
#### 3 Missing imputation
##### 3.1 Missing inspection
##### 3.2 Filtering
# By feature
feature_filter <- 
  filter_by_missing( data, 
                     method = 'feature',                        #feature or sample
                     threshold = params$filter_by_missing_feature_pct  #Percentage
  )

data <- feature_filter$x
features <- 
  features[feature_filter$index, ] %>%
  left_join(f.info, by = "Identity")

```

### Internal standard
```{r, eval = params$run.pos}
# IS by feature
if (nrow(peaks.is > 0)){

  feature_filter.is <- 
    filter_by_missing( data.is, 
                       method = 'feature',                        #feature or sample
                       threshold = params$filter_by_missing_feature_pct  #Percentage
    )
  
  data.is <- feature_filter.is$x
  features.is <- 
    features.is[feature_filter.is$index, ] %>%
    left_join(f.info, by = "Identity")
}
```

## 1.4 Missing distribution {.tabset .tabset-pills}
```{r, eval = params$run.pos}

# index.keep <- 
#   gsub(".mzML Peak area", "", colnames(data)) %in% c("Identity", params$sample.info.pos$Sample.name)
  
data_sub <- data

df <- 
  data_sub %>%
  rownames_to_column("Sample") %>% 
  pivot_longer(-Sample, names_to = "Feature", values_to = "var") %>%
  mutate(isna = (is.na(var)|var<=0))

p1 <- ggplot(df, aes(x = Sample, y = Feature, fill = isna)) +
  geom_raster() +
  scale_fill_manual(name = "",
                    values = c('#132b43', '#4fa3e4'),
                    labels = c("Present", "Missing")) +
  labs(title = "Missing values") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

data.plot <- data_sub
miss.sample <- data.frame(Sample = rownames(data.plot),
                          missing = apply(data.plot, 1, function(x){sum(is.na(x)|x<=0)/length(x)*100}))


p2 <- ggplot(miss.sample, aes(x = Sample,y = missing)) +
  geom_bar(stat = "identity") +
  labs( x = "Samples",
        y = "Proportion of missing values (%)") +
  theme_bw() +
  theme(axis.text.x = element_blank())
  
miss.feature <- data.frame(Feature = colnames(data.plot),
                           missing = apply(data.plot, 2, function(x){sum(is.na(x)|x<=0)/length(x)*100}))

p3 <- ggplot(miss.feature, aes(x = Feature,y = missing)) +
  geom_bar(stat = "identity") +
  labs( x = "Features",
        y = "Proportion of missing values (%)") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

### Overall
```{r, eval = params$run.pos, fig.width = 11, fig.height = 5}
p1
```

### Missing in samples
```{r, eval = params$run.pos, fig.width = 9, fig.height = 4}
ggplotly(p2)
```

### Missing in features
```{r, eval = params$run.pos, fig.width = 9, fig.height = 4}
ggplotly(p3)
```

## 1.5 Remove duplicate internal standard
```{r remove dup is, eval = params$run.pos}
#### 4 Remove duplicate internal standard
if (nrow(peaks.is > 0)){

  features.is.keep <-
    features.is %>%
    rowwise() %>%
    mutate(Identity_new = strsplit(Identity, "-iso")[[1]][1]) %>%
    group_by(Identity_new) %>%
    arrange(Identity_new, desc(get(make.names(paste0("mean.", params$po.sample.to.use))))) %>%
    filter(row_number() == 1) %>%
    ungroup() %>%
    select(-Identity_new)
  
  data.is.keep <-
    data.is[, features.is.keep$Identity]
}

print( "The following internal standards have been removed:", quote = FALSE)
cat(features.is$Identity[!features.is$Identity %in% features.is.keep$Identity], sep = '\n')
```

## 1.6 Export
```{r export, eval = params$run.pos}

### Prepare feature info

peaktable.cleaned <-
  cbind( features[, c("mz", "rt", "Identity", names(f.info)[-1])], 
         t(data)) %>%
  rowwise() %>% 
  mutate( Identity = strsplit( Identity, "-iso")[[1]][1] ) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  select( -n )

features$Identity_orig <- features$Identity
features$Identity <- peaktable.cleaned$Identity

if (nrow(peaks.is > 0)){

  peaktable.is.cleaned <-
    cbind( features.is.keep[, c("mz", "rt", "Identity", names(f.info)[-1])], 
           t(data.is.keep)) %>%
    rowwise() %>% 
    mutate( Identity = strsplit( Identity, "-iso")[[1]][1] ) %>%
    group_by(Identity) %>%
    mutate( n = n(), 
            Identity = ifelse( n > 1, 
                               paste0(Identity, "-iso", row_number()),
                               Identity)) %>%
    ungroup() %>%
    select( -n )
  
  features.is.keep$Identity_orig <- features.is.keep$Identity
  features.is.keep$Identity <- peaktable.is.cleaned$Identity
  
} else {

  peaktable.is.cleaned <- data.frame()
  features.is.keep <- data.frame()
}

peaktable.cleaned <-
  bind_rows(peaktable.cleaned, peaktable.is.cleaned)
        
output.file <- paste0(path.output, "peaktable_pos_cleaned.xlsx")

features.cleaned <-
  bind_rows(features, features.is.keep)
  
wb <- createWorkbook()
addWorksheet(wb, "peaktable")
writeData(wb, "peaktable", peaktable.cleaned)
addWorksheet(wb, "features")
writeData(wb, "features", features.cleaned)
saveWorkbook(wb, file = output.file, overwrite = TRUE)

print( paste0("POS cleaned peaktable was saved as: ", output.file ), quote = FALSE)

data_clean_pos <- list(data = peaktable.cleaned,
                       features = features.cleaned)
saveRDS(data_clean_pos, file = paste0(params$path.result, "/data/data_clean_pos.rds"))
```

$~$

# 2 Negative mode
## 2.1 Data import
```{r data import neg, eval = params$run.neg}

if (is.data.frame(params$path.peaktable.neg)){
  
  peaktable <- params$path.peaktable.neg
  
} else if ( grepl(".xlsx", params$path.peaktable.neg) ){
  
  peaktable <-
    readxl::read_excel( params$path.peaktable.neg )
  
} else {
  
  peaktable <-
    read.csv( params$path.peaktable.neg, 
              sep = params$peaktable.sep, 
              check.names = FALSE)
  
}

print("NEG peaktable was imported from the following file:", quote = FALSE)
print(params$path.peaktable.neg, quote = FALSE)

peaks.names <-
  paste0(params$sample.info.neg$Sample.name, params$peaktable.sample.pattern)
  
peaks.sample <-
  peaktable[, peaks.names] %>%
  mutate_if(is.character, as.numeric)

peaks.all <-
  cbind(peaktable[, c(params$mz_col_nr, params$rt_col_nr, params$identity_col_nr)], peaks.sample)

f.info <- 
  peaktable[, c(params$identity_col_nr, params$mz_col_nr, params$rt_col_nr, params$others), drop = FALSE] %>%
  dplyr::rename( Identity = 1,
                 mz = 2,
                 rt = 3) %>%
  mutate(  Identity = ifelse( grepl(": 0.", Identity, fixed = TRUE),
                               substr(Identity, 1, regexpr("\\:[^\\:]*$", Identity)-1),
                               Identity),
           Identity = ifelse( is.na(Identity),
                             paste0("MZ_", round(mz, 3), "_RT_", round(rt, 3)),
                             Identity ),
          Feature.id = row_number()) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  arrange(Feature.id) %>%
  select( -c(n, Feature.id, mz, rt) )

peaks.temp <-
  peaks.all %>%
  dplyr::rename( mz = 1,
                 rt = 2,
                 Identity = 3) %>%
  mutate( Identity = ifelse( grepl(": 0.", Identity, fixed = TRUE),
                               substr(Identity, 1, regexpr("\\:[^\\:]*$", Identity)-1),
                               Identity),
          Identity = ifelse( is.na(Identity),
                             paste0("MZ_", round(mz, 3), "_RT_", round(rt, 3)),
                             Identity ),
          Feature.id = row_number()) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  arrange(Feature.id) %>%
  select( -c(n, Feature.id) ) %>%
  mutate( Identity = ifelse(nchar(Identity) > 1000, substr(Identity, 1, 1000), Identity)) 

peaks.is <-
  peaks.temp %>%
  filter(grepl("^[0-9][0-9]\\.", substr(Identity, 1, 3)))

peaks <-
  peaks.temp %>%
  filter(!grepl("^[0-9][0-9]\\.", substr(Identity, 1, 3)))

datalist <- list(peaks.raw = peaks,
                 peaks.is.raw = peaks.is)
```

## 2.2 Feature filtering
```{r filtering neg, eval = params$run.neg}

# Peaks
mzrt.filtered <-
  filter_peaks( peaktable = peaks,
                sample.type = params$sample.info.neg$Sample.type,
                rsd.filter = params$rsd.filter,
                rt.range = params$rt.range,
                mean.filter = params$mean.filter)

peaks.filtered <- 
  mzrt.filtered$peaktable

features <- 
  cbind( peaks.filtered[, c("mz", "rt", "Identity")], mzrt.filtered$summary[, -1])

data <- 
  t(peaks.filtered[, !colnames(peaks.filtered) %in% c("mz", "rt", "Identity")]) %>%
  as.data.frame()

colnames(data) <- features$Identity

# IS
if (nrow(peaks.is > 0)){

  mzrt.is.filtered <-
    filter_peaks( peaktable = peaks.is,
                  sample.type = params$sample.info.neg$Sample.type)
  
  peaks.is.filtered <- 
    mzrt.is.filtered$peaktable
  
  
  features.is <- 
    cbind( peaks.is.filtered[, c("mz", "rt", "Identity")], mzrt.is.filtered$summary[, -1])
  
  data.is <- 
    t(peaks.is.filtered[, !colnames(peaks.is.filtered) %in% c("mz", "rt", "Identity")]) %>%
    as.data.frame()
  
  colnames(data.is) <- features.is$Identity
}
```

## 2.3 Filter based on missings
### Feature
```{r filter neg, eval = params$run.neg}

# By feature
feature_filter <- 
  filter_by_missing( data, 
                     method = 'feature',                        #feature or sample
                     threshold = params$filter_by_missing_feature_pct  #Percentage
  )

data <- feature_filter$x
features <- 
  features[feature_filter$index, ] %>%
  left_join(f.info, by = "Identity")

```

### Internal standard
```{r, eval = params$run.neg}
# IS by feature
if (nrow(peaks.is > 0)){

  feature_filter.is <- 
    filter_by_missing( data.is, 
                       method = 'feature',                        #feature or sample
                       threshold = params$filter_by_missing_feature_pct  #Percentage
    )
  
  data.is <- feature_filter.is$x
  features.is <- 
    features.is[feature_filter.is$index, ] %>%
    left_join(f.info, by = "Identity")
}
```

## 2.4 Missing distribution {.tabset .tabset-pills} 
```{r, eval = params$run.neg}

# index.keep <- 
#   gsub(".mzML Peak area", "", colnames(data)) %in% c("Identity", sample.info.neg$Sample.name)
  
data_sub <- data

df <- 
  data_sub %>%
  rownames_to_column("Sample") %>% 
  pivot_longer(-Sample, names_to = "Feature", values_to = "var") %>%
  mutate(isna = (is.na(var)|var<=0))

p1 <- ggplot(df, aes(x = Sample, y = Feature, fill = isna)) +
  geom_raster() +
  scale_fill_manual(name = "",
                    values = c('#132b43', '#4fa3e4'),
                    labels = c("Present", "Missing")) +
  labs(title = "Missing values") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

data.plot <- data_sub
miss.sample <- data.frame(Sample = rownames(data.plot),
                          missing = apply(data.plot, 1, function(x){sum(is.na(x)|x<=0)/length(x)*100}))


p2 <- ggplot(miss.sample, aes(x = Sample,y = missing)) +
  geom_bar(stat = "identity") +
  labs( x = "Samples",
        y = "Proportion of missing values (%)") +
  theme_bw() +
  theme(axis.text.x = element_blank())
  
miss.feature <- data.frame(Feature = colnames(data.plot),
                           missing = apply(data.plot, 2, function(x){sum(is.na(x)|x<=0)/length(x)*100}))

p3 <- ggplot(miss.feature, aes(x = Feature,y = missing)) +
  geom_bar(stat = "identity") +
  labs( x = "Features",
        y = "Proportion of missing values (%)") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

### Overall
```{r, eval = params$run.neg, fig.width = 11, fig.height = 5}
p1
```

### Missing in samples
```{r, eval = params$run.neg, fig.width = 9, fig.height = 4}
ggplotly(p2)
```

### Missing in features
```{r, eval = params$run.neg, fig.width = 9, fig.height = 4}
ggplotly(p3)
```

$~$

## 2.5 Remove duplicate internal standard
```{r remove dup is neg, eval = params$run.neg}

if (nrow(peaks.is > 0)){

  features.is.keep <-
    features.is %>%
    rowwise() %>%
    mutate(Identity_new = strsplit(Identity, "-iso")[[1]][1]) %>%
    group_by(Identity_new) %>%
    arrange(Identity_new, desc(get(make.names(paste0("mean.", params$po.sample.to.use))))) %>%
    filter(row_number() == 1) %>%
    ungroup() %>%
    select(-Identity_new)
  
  data.is.keep <-
    data.is[, features.is.keep$Identity]
}

print( "The following internal standards have been removed:", quote = FALSE)
cat(features.is$Identity[!features.is$Identity %in% features.is.keep$Identity], sep = '\n')

```

## 2.6 Export
```{r export neg, eval = params$run.neg}

### Prepare feature info

peaktable.cleaned <-
  cbind( features[, c("mz", "rt", "Identity", names(f.info)[-1]),], 
         t(data)) %>%
  rowwise() %>% 
  mutate( Identity = strsplit( Identity, "-iso")[[1]][1] ) %>%
  group_by(Identity) %>%
  mutate( n = n(), 
          Identity = ifelse( n > 1, 
                             paste0(Identity, "-iso", row_number()),
                             Identity)) %>%
  ungroup() %>%
  select( -n )

features$Identity_orig <- features$Identity
features$Identity <- peaktable.cleaned$Identity

if (nrow(peaks.is > 0)){

  peaktable.is.cleaned <-
    cbind( features.is.keep[, c("mz", "rt", "Identity", names(f.info)[-1])], 
           t(data.is.keep)) %>%
    rowwise() %>% 
    mutate( Identity = strsplit( Identity, "-iso")[[1]][1] ) %>%
    group_by(Identity) %>%
    mutate( n = n(), 
            Identity = ifelse( n > 1, 
                               paste0(Identity, "-iso", row_number()),
                               Identity)) %>%
    ungroup() %>%
    select( -n )
  
  features.is.keep$Identity_orig <- features.is.keep$Identity
  features.is.keep$Identity <- peaktable.is.cleaned$Identity

} else {
  peaktable.is.cleaned <- data.frame()
  features.is.keep <- data.frame()
}

peaktable.cleaned <-
  bind_rows(peaktable.cleaned, peaktable.is.cleaned)

output.file <- paste0(path.output, "peaktable_neg_cleaned.xlsx")
  
features.cleaned <-
  bind_rows(features, features.is.keep)

wb <- createWorkbook()
addWorksheet(wb, "peaktable")
writeData(wb, "peaktable", peaktable.cleaned)
addWorksheet(wb, "features")
writeData(wb, "features", features.cleaned)
saveWorkbook(wb, file = output.file, overwrite = TRUE)

print( paste0("NEG cleaned peaktable was saved as: ", output.file ), quote = FALSE)

data_clean_neg <- list(data = peaktable.cleaned,
                       features = features.cleaned)
saveRDS(data_clean_neg, file = paste0(params$path.result, "/data/data_clean_neg.rds"))
```
