---
title: "Normalization comparison"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"

knit: ( function(inputFile, encoding) { 
          rmarkdown::render( inputFile, 
                             encoding = encoding, 
                             output_file = paste0( substr(inputFile,1,nchar(inputFile)-4), '_', Sys.Date(),'.html')) 
          })
          
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: '4'
    df_print: paged
---

```{r setup environment, include = FALSE}

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

```

```{r set path}

path.output <- paste0(params$path.result, "06_Normalization_comparison/")

if (!dir.exists(path.output)) dir.create(path.output)

```

### POS
#### 1 Data import
```{r data import pos, eval = params$run.pos}

if (is.null(params$path.peaktable.pos)){
  
  params$path.peaktable.pos <- paste0(params$path.result, "05_Clean_peaktable/peaktable_targeted_pos_cleaned.xlsx")
  
}

print("POS peaktables were imported from the following files:", quote = FALSE)
print(params$path.peaktable.pos, quote = FALSE)
print( paste0("POS normalized peaktables were saved in: ", path.output ), quote = FALSE)

if (is.data.frame(params$path.peaktable.pos)){
  
  peaktable <- params$path.peaktable.pos
  
} else if ( grepl(".xlsx", params$path.peaktable.pos) ){
  
  peaktable <-
    readxl::read_excel( params$path.peaktable.pos ) 
  
} else {
  
  peaktable <-
    read.csv( params$path.peaktable.pos, 
              sep = params$peaktable.sep,
              check.names = FALSE)
  
}

if (!is.null(params$peaktable.sample.pattern)){
  
  f.names <- colnames(peaktable)[grepl(params$peaktable.sample.pattern, colnames(peaktable))]
  
} else {
  
  f.names <- params$sample_col_nr:ncol(peaktable)
  
}

f.info <- 
  peaktable[, c(params$mz_col_nr, params$rt_col_nr, params$identity_col_nr, params$others)]

colnames(f.info)[1:3] <- c("mz", "rt", "Identity")

peaks <-
  cbind(f.info, peaktable[, f.names])

datalist <-
  prepare_datalist( peaks = peaks,
                    sample.info = params$sample.info.pos,
                    sample.pattern = params$peaktable.sample.pattern,
                    type.to.remove = params$type.to.remove )

datalist$feature.info <-
  datalist$feature.info %>%
  left_join(f.info[, c(3:ncol(f.info))], by = "Identity")

```

#### 2 Filter based on missings
```{r filter pos, eval = params$run.pos}

datalist <-
  filter_sample_feature( datalist = datalist,
                         missing.sample.thres = params$missing.sample.thres,
                         missing.feature.thres = params$missing.feature.thres)

```

#### 3 Missing imputation
```{r impute pos, eval = params$run.pos}

datalist <-
  missing_impute( datalist = datalist,
                  missing.impute.method.sample = params$missing.impute.method.sample,
                  missing.impute.method.is = params$missing.impute.method.is)
    

```

#### 4 Outlier removal
```{r check data with PCA pos, eval = params$run.pos}

p.pca.raw <- 
  plot_PCA(
    datalist$data,
    group = as.factor(datalist$sample.info$Sample.type),
    alpha = 0.25,
    plotly.text = datalist$sample.info$Sample.id
  )

ggplotly(p.pca.raw)

```

```{r remove outliers pos, eval = params$run.pos}

datalist <-
  filter_sample_feature( datalist = datalist,
                         filter.sample.remove = params$outliers.sample,
                         filter.feature.remove = params$outliers.feature,
                         filter.is.remove = params$outliers.is)


```


#### 5 Check the intensities of internal standards {.tabset}
```{r explore internal standard pos, include = FALSE, eval = params$run.pos}

plt.is.rsd <-
  plot_istd_rsd(datalist = datalist)


```

##### RSD bar plot
```{r rsd.bar.plot pos, eval = params$run.pos}

ggplotly(plt.is.rsd$plt.rsd, width = 8)

```

##### IS intensities in samples
```{r rsd.sample pos, eval = params$run.pos}

plt.is.rsd$plt.int

```


#### 6 Normalization comparison
```{r normalization setup pos, include = FALSE, eval = params$run.pos}

process.list <-
  compute_normalization( 
    datalist = datalist,
    norm.method = params$norm.method,
    path.result = path.output,
    prefix = "pos_",
    batch.wise = FALSE,
    po.sample.to.use = params$po.sample.to.use,
    feature.info = datalist$feature.info)

```

```{r evaluate normalization pos, include = FALSE, eval = params$run.pos}

pca.plot <- 
  plot_list_pca(
    x = process.list,
    group = datalist$sample.info$Batch,
    group.label = "Batch",
    plotly.text = datalist$sample.info$Sample.id)

pca.plot.type <- 
  plot_list_pca(
    x = process.list,
    group = datalist$sample.info$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$sample.info$Sample.id)

plt.eval <-
  plot_eval( x = process.list,
             type = datalist$sample.info$Sample.type)

norm_pos <- list(#process.list = process.list,
                 pca.plot = pca.plot,
                  pca.plot.type = pca.plot.type,
                  plt.eval = plt.eval)

saveRDS(norm_pos, file = paste0(params$path.result, "/data/data_norm_pos.rds"))
```

##### PCA {.tabset}

###### Raw Data

```{r PCA-Loaded pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

a <- ggplotly(pca.plot$raw.data, session="knitr", showlegend = T)
b <- ggplotly(pca.plot.type$raw.data, session="knitr", showlegend = T)

subplot(a, b)
```

###### Best IS

```{r PCA-Bestid pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$bestis)){

  a <- ggplotly(pca.plot$bestis, session="knitr")
  b <- ggplotly(pca.plot.type$bestis, session="knitr")
  
  subplot(a, b)
}

```

###### Lowest CV

```{r PCA-Lowest CV pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$low_cv)){

  a <- ggplotly(pca.plot$low_cv, session="knitr")
  b <- ggplotly(pca.plot.type$low_cv, session="knitr")
  
  subplot(a, b)
}

```

###### NOMIS

```{r PCA-Normalized_by_NOMIS pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$nomis)){

  a <- ggplotly(pca.plot$nomis, session="knitr")
  b <- ggplotly(pca.plot.type$nomis, session="knitr")
  
  subplot(a, b)
}
```

###### Cubic splines

```{r PCA-Normalized_by_qspline pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$qspline)){

  a <- ggplotly(pca.plot$qspline, session="knitr")
  b <- ggplotly(pca.plot.type$qspline, session="knitr")
  
  subplot(a, b)
}

```

###### Loess

```{r PCA-Normalized_by_loess pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$loess)){
  a <- ggplotly(pca.plot$loess, session="knitr")
  b <- ggplotly(pca.plot.type$loess, session="knitr")
  
  subplot(a, b)
}
```

###### Combat

```{r PCA-Normalized_by_combat pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$combat)){

  a <- ggplotly(pca.plot$combat, session="knitr")
  b <- ggplotly(pca.plot.type$combat, session="knitr")
  
  subplot(a, b)
}

```

###### limma

```{r PCA-Normalized_by_limma pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$limma)){

  a <- ggplotly(pca.plot$limma, session="knitr")
  b <- ggplotly(pca.plot.type$limma, session="knitr")
  
  subplot(a, b)
}

```

###### Sum

```{r PCA-Normalized_by_sum pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$sum)){

  a <- ggplotly(pca.plot$sum, session="knitr")
  b <- ggplotly(pca.plot.type$sum, session="knitr")
  
  subplot(a, b)
}
```

###### Median

```{r PCA-Normalized_by_median pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$median)){

  a <- ggplotly(pca.plot$median, session="knitr")
  b <- ggplotly(pca.plot.type$median, session="knitr")
  
  subplot(a, b)
}
```

###### PQN

```{r PCA-Normalized_by_PQN pos, fig.height = 4, fig.width = 8, eval = params$run.pos}

if (!is.null(pca.plot$pqn)){

  a <- ggplotly(pca.plot$pqn, session="knitr")
  b <- ggplotly(pca.plot.type$pqn, session="knitr")
  
  subplot(a, b)
}
```

##### RSD for all features in QC samples {.tabset}

###### RSD
```{r RSD pos, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.pos}

plotly::ggplotly(plt.eval$plt.rsd, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.rsd.mean, session="knitr") %>%
  layout(boxmode = "group")

```

###### MAD
```{r MAD pos, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.pos}

plotly::ggplotly(plt.eval$plt.mad, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.mad.mean, session="knitr") %>%
  layout(boxmode = "group")

```

###### Variance
```{r VAR pos, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.pos}

plotly::ggplotly(plt.eval$plt.var, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.var.sum, session="knitr") %>%
  layout(boxmode = "group")

```


### NEG
#### 1 Data import
```{r data import neg, eval = params$run.neg}

if (is.null(params$path.peaktable.neg)){
  
  params$path.peaktable.neg <- paste0(params$path.result, "05_Clean_peaktable/peaktable_targeted_neg_cleaned.xlsx")
  
}

print("NEG peaktables were imported from the following files:", quote = FALSE)
print(params$path.peaktable.neg, quote = FALSE)
print( paste0("POS normalized peaktables were saved in: ", path.output ), quote = FALSE)

if (is.data.frame(params$path.peaktable.neg)){
  
  peaktable <- params$path.peaktable.neg
  
} else if ( grepl(".xlsx", params$path.peaktable.neg) ){
  
  peaktable <-
    readxl::read_excel( params$path.peaktable.neg ) 
  
} else {
  
  peaktable <-
    read.csv( params$path.peaktable.neg, 
              sep = params$peaktable.sep,
              check.names = FALSE)
  
}

if (!is.null(params$peaktable.sample.pattern)){
  
  f.names <- colnames(peaktable)[grepl(params$peaktable.sample.pattern, colnames(peaktable))]
  
} else {
  
  f.names <- params$sample_col_nr:ncol(peaktable)
  
}

f.info <- 
  peaktable[, c(params$mz_col_nr, params$rt_col_nr, params$identity_col_nr, params$others)]

colnames(f.info)[1:3] <- c("mz", "rt", "Identity")

peaks <-
  cbind(f.info, peaktable[, f.names])

datalist <-
  prepare_datalist( peaks = peaks,
                    sample.info = params$sample.info.neg,
                    sample.pattern = params$peaktable.sample.pattern,
                    type.to.remove = params$type.to.remove )

datalist$feature.info <-
  datalist$feature.info %>%
  left_join(f.info[, c(3:ncol(f.info))], by = "Identity")

```

#### 2 Filter based on missings
```{r filter neg, eval = params$run.neg}

datalist <-
  filter_sample_feature( datalist = datalist,
                         missing.sample.thres = params$missing.sample.thres,
                         missing.feature.thres = params$missing.feature.thres)

```

#### 3 Missing imputation
```{r impute neg, eval = params$run.neg}

datalist <-
  missing_impute( datalist = datalist,
                  missing.impute.method.sample = params$missing.impute.method.sample,
                  missing.impute.method.is = params$missing.impute.method.is)
    

```

#### 4 Outlier removal
```{r check data with PCA neg, eval = params$run.neg}

p.pca.raw <- 
  plot_PCA(
    datalist$data,
    group = as.factor(datalist$sample.info$Sample.type),
    alpha = 0.25,
    plotly.text = datalist$sample.info$Sample.id
  )

ggplotly(p.pca.raw)

```

```{r remove outliers neg, eval = params$run.neg}

datalist <-
  filter_sample_feature( datalist = datalist,
                         filter.sample.remove = params$outliers.sample,
                         filter.feature.remove = params$outliers.feature,
                         filter.is.remove = params$outliers.is)


```


#### 5 Check the intensities of internal standards {.tabset}
```{r explore internal standard neg, include = FALSE, eval = params$run.neg}

plt.is.rsd <-
  plot_istd_rsd(datalist = datalist)


```

##### RSD bar plot
```{r rsd.bar.plot neg, eval = params$run.neg}

ggplotly(plt.is.rsd$plt.rsd, width = 8)

```

##### IS intensities in samples
```{r rsd.sample neg, eval = params$run.neg}

plt.is.rsd$plt.int

```


#### 6 Normalization comparison
```{r normalization setup neg, include = FALSE, eval = params$run.neg}

process.list <-
  compute_normalization( 
    datalist = datalist,
    norm.method = params$norm.method,
    path.result = path.output,
    prefix = "neg_",
    batch.wise = FALSE,
    po.sample.to.use = params$po.sample.to.use,
    feature.info = datalist$feature.info)

```

```{r evaluate normalization neg, include = FALSE, eval = params$run.neg}

pca.plot <- 
  plot_list_pca(
    x = process.list,
    group = datalist$sample.info$Batch,
    group.label = "Batch",
    plotly.text = datalist$sample.info$Sample.id)

pca.plot.type <- 
  plot_list_pca(
    x = process.list,
    group = datalist$sample.info$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$sample.info$Sample.id)

plt.eval <-
  plot_eval( x = process.list,
             type = datalist$sample.info$Sample.type)

norm_neg <- list(#process.list = process.list,
                 pca.plot = pca.plot,
                  pca.plot.type = pca.plot.type,
                  plt.eval = plt.eval)

saveRDS(norm_neg, file = paste0(params$path.result, "/data/data_norm_neg.rds"))
```

##### PCA {.tabset}

###### Raw Data

```{r PCA-Loaded neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

a <- ggplotly(pca.plot$raw.data, session="knitr", showlegend = T)
b <- ggplotly(pca.plot.type$raw.data, session="knitr", showlegend = T)

subplot(a, b)
```

###### Best IS

```{r PCA-Bestid neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$bestis)){

  a <- ggplotly(pca.plot$bestis, session="knitr")
  b <- ggplotly(pca.plot.type$bestis, session="knitr")
  
  subplot(a, b)
}

```

###### Lowest CV

```{r PCA-Lowest CV neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$low_cv)){

  a <- ggplotly(pca.plot$low_cv, session="knitr")
  b <- ggplotly(pca.plot.type$low_cv, session="knitr")
  
  subplot(a, b)
}

```

###### NOMIS

```{r PCA-Normalized_by_NOMIS neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$nomis)){

  a <- ggplotly(pca.plot$nomis, session="knitr")
  b <- ggplotly(pca.plot.type$nomis, session="knitr")
  
  subplot(a, b)
}
```

###### Cubic splines

```{r PCA-Normalized_by_qspline neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$qspline)){

  a <- ggplotly(pca.plot$qspline, session="knitr")
  b <- ggplotly(pca.plot.type$qspline, session="knitr")
  
  subplot(a, b)
}

```

###### Loess

```{r PCA-Normalized_by_loess neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$loess)){
  a <- ggplotly(pca.plot$loess, session="knitr")
  b <- ggplotly(pca.plot.type$loess, session="knitr")
  
  subplot(a, b)
}
```

###### Combat

```{r PCA-Normalized_by_combat neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$combat)){

  a <- ggplotly(pca.plot$combat, session="knitr")
  b <- ggplotly(pca.plot.type$combat, session="knitr")
  
  subplot(a, b)
}

```

###### limma

```{r PCA-Normalized_by_limma neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$limma)){

  a <- ggplotly(pca.plot$limma, session="knitr")
  b <- ggplotly(pca.plot.type$limma, session="knitr")
  
  subplot(a, b)
}

```

###### Sum

```{r PCA-Normalized_by_sum neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$sum)){

  a <- ggplotly(pca.plot$sum, session="knitr")
  b <- ggplotly(pca.plot.type$sum, session="knitr")
  
  subplot(a, b)
}
```

###### Median

```{r PCA-Normalized_by_median neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$median)){

  a <- ggplotly(pca.plot$median, session="knitr")
  b <- ggplotly(pca.plot.type$median, session="knitr")
  
  subplot(a, b)
}
```

###### PQN

```{r PCA-Normalized_by_PQN neg, fig.height = 4, fig.width = 8, eval = params$run.neg}

if (!is.null(pca.plot$pqn)){

  a <- ggplotly(pca.plot$pqn, session="knitr")
  b <- ggplotly(pca.plot.type$pqn, session="knitr")
  
  subplot(a, b)
}
```

##### RSD for all features in QC samples {.tabset}

###### RSD
```{r RSD neg, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.neg}

plotly::ggplotly(plt.eval$plt.rsd, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.rsd.mean, session="knitr") %>%
  layout(boxmode = "group")

```

###### MAD
```{r MAD neg, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.neg}

plotly::ggplotly(plt.eval$plt.mad, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.mad.mean, session="knitr") %>%
  layout(boxmode = "group")

```

###### Variance
```{r VAR neg, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, eval = params$run.neg}

plotly::ggplotly(plt.eval$plt.var, session="knitr") %>%
  layout(boxmode = "group")

plotly::ggplotly(plt.eval$plt.var.sum, session="knitr") %>%
  layout(boxmode = "group")

```

```{r SessionInfo, include = FALSE}

# SessionInfo

utils::sessionInfo()

```
