---
title: "QC internal standards"
author: "`r params$author`"
date: "`r base::Sys.Date()`"

knit: ( function(inputFile, encoding) { 
          rmarkdown::render( inputFile, 
                             encoding = encoding, 
                             output_file = paste0( substr(inputFile,1,nchar(inputFile)-4), '_', Sys.Date(),'.html')) 
          })
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '4'
    df_print: paged
---

<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE}

library(kableExtra)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

```

# 1 Data import
```{r import data}

if (is.null(params$path.peaktable.IS.pos)){
  
  params$path.peaktable.IS.pos <- paste0(params$path.result, "01_01_Preprocessing_IS/peaktable_targeted_IS_pos.csv")

}
print( paste0("POS peaktable for QC was imported from: ", params$path.peaktable.IS.pos), quote = FALSE)

if (is.null(params$path.peaktable.IS.neg)){
  
  params$path.peaktable.IS.neg <- paste0(params$path.result, "01_01_Preprocessing_IS/peaktable_targeted_IS_neg.csv")

}
print( paste0("NEG peaktable for QC was imported from: ", params$path.peaktable.IS.neg), quote = FALSE)

if (is.null(params$sample.info.pos.qc)){
  params$sample.info.pos.qc <- params$sample.info.pos
}

if (is.null(params$sample.info.neg.qc)){
  params$sample.info.neg.qc <- params$sample.info.neg
}

```

$~$

# 2 QC of internal standards
## 2.1 Positive mode
```{r evaluation pos, include = FALSE, eval = params$run.pos}

# Extract theoretical and actual internal standard
combined_file <-
  extract_is_other( sample.info = params$sample.info.pos.qc,
              reference.type = params$reference.type,
              lib.istd = params$lib.is[[params$method]]$pos,
              path.peaktable.IS = params$path.peaktable.IS.pos,
              path.result = params$path.result) %>% 
  filter(!ISTD %in% params$istd.remove.pos) %>% 
  mutate(ISTD = gsub("-iso1", "", ISTD))

plt <- create_qc(sampled_data = combined_file,
                 figure_height = params$figure_height,
                 figure_width = params$figure_width)

summary <- 
  combined_file %>%
  select(ISTD, Theoretical.rt, median.rt, Theoretical.mz, median.mz, median.intensity) %>%
  unique() %>% 
  mutate(rt.diff = median.rt - Theoretical.rt,
         mz.diff = median.mz - Theoretical.mz) %>% 
  relocate(rt.diff, .after = median.rt) %>% 
  relocate(mz.diff, .after = median.mz)

qc_is_pos <- list(combined_file = combined_file,
                  plt = plt,
                  summary = summary)

saveRDS(qc_is_pos, file = paste0(params$path.result, "/data/QC_IS_pos.rds"))
```

### 2.1.1 Summary {.tabset .tabset-pills}
#### Internal standard
```{r ms overall pos, eval = params$run.pos}
summary %>%
  mutate_if(is.numeric, function(x){round(x, 2)}) %>% 
  dplyr::rename("Internal standard" = "ISTD",            
                "Theoretical RT" = "Theoretical.rt",
                "Median RT" = "median.rt",        
                "RT difference" = "rt.diff",          
                "Theoretical m/z" = "Theoretical.mz",
                "Median m/z" = "median.mz",
                "m/z difference" = "mz.diff",          
                "Median intensity" = "median.intensity") %>% 
  kbl() %>%
  kable_paper()
```

#### RSD
```{r, eval = params$run.pos, fig.width = 8, fig.height = 4}
plt[[7]]
```

#### Shift
```{r ms overall pos2, eval = params$run.pos, fig.width = 8, fig.height = 4}
p1 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=rt.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color = "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=rt.diff), color = "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in retention time")

p2 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=mz.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color= "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=mz.diff), color= "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in m/z")

subplot(p1, p2, nrows = 1, margin = 0.04, titleX = TRUE, titleY = TRUE, shareY = TRUE) 

```

### 2.1.2 QC of individual samples {.tabset .tabset-pills}
#### Retention time
```{r ms RT pos, eval = params$run.pos, fig.width = 10, fig.height = 6}

subplot(plt[[1]], plt[[2]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

#### m/z
```{r ms m/z pos, eval = params$run.pos, fig.width = 10, fig.height = 6}

subplot(plt[[3]], plt[[4]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

#### Intensity
```{r ms1p Int pos, eval = params$run.pos, fig.width = 10, fig.height = 6}

subplot(plt[[5]], plt[[6]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

## 2.2 Negative mode
```{r evaluation neg, include = FALSE, eval = params$run.neg}

# Extract theoretical and actual internal standard
combined_file <-
  extract_is_other( sample.info = params$sample.info.neg.qc,
              reference.type = params$reference.type,
              lib.istd = params$lib.is[[params$method]]$neg,
              path.peaktable.IS = params$path.peaktable.IS.neg,
              path.result = params$path.result) %>% 
  filter(!ISTD %in% params$istd.remove.neg) %>% 
  mutate(ISTD = gsub("-iso1", "", ISTD))

plt <- create_qc(sampled_data = combined_file,
                 figure_height = params$figure_height,
                 figure_width = params$figure_width)

summary <- 
  combined_file %>%
  select(ISTD, Theoretical.rt, median.rt, Theoretical.mz, median.mz, median.intensity) %>%
  unique() %>% 
  mutate(rt.diff = median.rt - Theoretical.rt,
         mz.diff = median.mz - Theoretical.mz) %>% 
  relocate(rt.diff, .after = median.rt) %>% 
  relocate(mz.diff, .after = median.mz)

qc_is_neg <- list(combined_file = combined_file,
                  plt = plt,
                  summary = summary)

saveRDS(qc_is_neg, file = paste0(params$path.result, "/data/QC_IS_neg.rds"))
```

### 2.2.1 Summary {.tabset .tabset-pills}
#### Internal standard
```{r ms overall neg, eval = params$run.neg}
summary %>%
  mutate_if(is.numeric, function(x){round(x, 2)}) %>% 
  dplyr::rename("Internal standard" = "ISTD",            
                "Theoretical RT" = "Theoretical.rt",
                "Median RT" = "median.rt",        
                "RT difference" = "rt.diff",          
                "Theoretical m/z" = "Theoretical.mz",
                "Median m/z" = "median.mz",
                "m/z difference" = "mz.diff",          
                "Median intensity" = "median.intensity") %>% 
  kbl() %>%
  kable_paper()
```

#### RSD
```{r, eval = params$run.neg, fig.width = 8, fig.height = 4}
plt[[7]]
```

#### Shift
```{r ms overall neg2, eval = params$run.neg, fig.width = 8, fig.height = 4}
p1 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=rt.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color = "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=rt.diff), color = "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in retention time")

p2 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=mz.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color= "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=mz.diff), color= "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in m/z")

subplot(p1, p2, nrows = 1, margin = 0.04, titleX = TRUE, titleY = TRUE, shareY = TRUE) 

```

### 2.2.2 QC of individual samples {.tabset .tabset-pills}
#### Retention time
```{r ms RT neg, eval = params$run.neg, fig.width = 10, fig.height = 6}

subplot(plt[[1]], plt[[2]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

#### m/z
```{r ms m/z neg, eval = params$run.neg, fig.width = 10, fig.height = 6}

subplot(plt[[3]], plt[[4]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

#### Intensity
```{r ms1p Int neg, eval = params$run.neg, fig.width = 10, fig.height = 6}

subplot(plt[[5]], plt[[6]], nrows = 2, margin = 0.04, titleX = TRUE, titleY = TRUE) 

```

