---
title: "QC report"
author: "`r params$author`"
date: "`r base::Sys.Date()`"

knit: ( function(inputFile, encoding) { 
          rmarkdown::render( inputFile, 
                             encoding = encoding, 
                             output_file = paste0( substr(inputFile,1,nchar(inputFile)-4), '_', Sys.Date(),'.html')) 
          })
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '4'
    df_print: paged
---

<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 11px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE}

library(metpipe)
library(ggplot2)
library(openxlsx)
library(tidyverse)
library(kableExtra)
library(ggsci)
library(plotly)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

```

```{r import data}

path.data <- paste0(params$path.result, "/data/")
# params$figure_height <- 500
# params$figure_width <- 1000

### QC_IS
qc_is <- list()
#data_clean <- list()
#data_norm <- list()
data_merge <- readRDS(paste0(params$path.result, "data/data_merge.rds"))

if (params$run.pos){
  qc_is$pos <- readRDS(paste0(params$path.result, "data/QC_IS_pos.rds"))
  #data_clean$pos <- readRDS(paste0(params$path.result, "data/data_clean_pos.rds"))
  #data_norm$pos <- readRDS(paste0(params$path.result, "data/data_norm_pos.rds"))
  
  qc_is$pos$plt <- create_qc(sampled_data = qc_is$pos$combined_file,
                             figure_height = params$figure_height,
                             figure_width = params$figure_width)
  
  sample.info.pos <- 
    params$sample.info.pos %>% 
    filter(Sample.type == "Sample")
}

if (params$run.neg){
  qc_is$neg <- readRDS(paste0(params$path.result, "data/QC_IS_neg.rds"))
  #data_clean$neg <- readRDS(paste0(params$path.result, "data/data_clean_neg.rds"))
  #data_norm$neg <- readRDS(paste0(params$path.result, "data/data_norm_neg.rds"))
  
  qc_is$neg$plt <- create_qc(sampled_data = qc_is$neg$combined_file,
                             figure_height = params$figure_height,
                             figure_width = params$figure_width)
  
  sample.info.neg <- 
    params$sample.info.neg %>% 
    filter(Sample.type == "Sample")
}

```

$~$

# 1 Sample overview
```{r data overview}

metatbl <- 
  data_merge$sample.info %>% 
  filter(Sample.type == "Sample") %>% 
  select(Project, Sample) %>% 
  rownames_to_column("ID") %>% 
  mutate(Method = params$method,
         POS = Sample %in% sample.info.pos$Sample,
         NEG = Sample %in% sample.info.neg$Sample,
         Peaktable = POS & NEG) # change later with merged data

# knitr::kable(metatbl)
kbl(metatbl) %>%
  kable_paper() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>% 
  scroll_box(width = "900px", height = "400px") %>%
  column_spec(5, color = "white",
              background = spec_color(metatbl$POS, end = 1, direction = -1)) %>%
  column_spec(6, color = "white",
              background = spec_color(metatbl$NEG, end = 1, direction = -1)) %>%
  column_spec(7, color = "white",
              background = spec_color(metatbl$Peaktable, end = 1, direction = -1))
```

$~$

# 2 QC of internal standards
## 2.1 Positive mode {.tabset .tabset-pills}
### Internal standard
```{r ms overall pos, eval = params$run.pos}
summary <- qc_is$pos$summary
plt <- qc_is$pos$plt

summary %>%
  mutate_if(is.numeric, function(x){round(x, 2)}) %>% 
  dplyr::rename("Internal standard" = "ISTD",            
                "Theoretical RT" = "Theoretical.rt",
                "Median RT" = "median.rt",        
                "RT difference" = "rt.diff",          
                "Theoretical m/z" = "Theoretical.mz",
                "Median m/z" = "median.mz",
                "m/z difference" = "mz.diff",          
                "Median intensity" = "median.intensity") %>% 
  kbl() %>%
  kable_paper()
```

### RSD
```{r, eval = params$run.pos, fig.height = 4, fig.width = 8}
plt[[7]]
```

### Shift
```{r ms overall pos2, eval = params$run.pos, fig.height = 4, fig.width = 8}
p1 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=rt.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color = "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=rt.diff), color = "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in retention time")

p2 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=mz.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color= "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=mz.diff), color= "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in m/z")

subplot(p1, p2, nrows = 1, margin = 0.04, titleX = TRUE, titleY = TRUE, shareY = TRUE) 

```

## 2.2 Negative mode {.tabset .tabset-pills}
### Internal standard
```{r ms overall neg, eval = params$run.neg}
summary <- qc_is$neg$summary
plt <- qc_is$neg$plt

summary %>%
  mutate_if(is.numeric, function(x){round(x, 2)}) %>% 
  dplyr::rename("Internal standard" = "ISTD",            
                "Theoretical RT" = "Theoretical.rt",
                "Median RT" = "median.rt",        
                "RT difference" = "rt.diff",          
                "Theoretical m/z" = "Theoretical.mz",
                "Median m/z" = "median.mz",
                "m/z difference" = "mz.diff",          
                "Median intensity" = "median.intensity") %>% 
  kbl() %>%
  kable_paper()
```

### RSD
```{r, eval = params$run.neg, fig.height = 4, fig.width = 8}
plt[[7]]
```

### Shift
```{r ms overall neg2, eval = params$run.neg, fig.height = 4, fig.width = 8}
p1 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=rt.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color = "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=rt.diff), color = "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in retention time")

p2 <- 
  ggplot(summary) +
    geom_segment( aes(x=ISTD, xend=ISTD, y=0, yend=mz.diff), color="grey") +
    geom_point( aes(x=ISTD, y=0), color= "#F39B7FB2", size=3 ) +
    geom_point( aes(x=ISTD, y=mz.diff), color= "#3C5488B2", size=3 ) +
    coord_flip()+
    scale_x_discrete(limits=rev) +
    theme_bw() +
    xlab("Internal standard") +
    ylab("Difference in m/z")

subplot(p1, p2, nrows = 1, margin = 0.04, titleX = TRUE, titleY = TRUE, shareY = TRUE) 

```

$~$

# 3 Batch evaluation {.tabset .tabset-pills}
## PCA
```{r, fig.height = 4, fig.width = 4}

data_all <- data_merge$data

p <- 
  plot_PCA(
      data_all[, -c(1:ncol(data_merge$sample.info))],
      group = data_all$Batch,
      group.label = "Batch",
      plotly.text = data_all$Sample,
      ratio = 1)
  
ggplotly(p)

```

## Sample mean across batches
```{r, fig.height = 4, fig.width = 8}

Sample.mean <- 
  apply(data_all[, -c(1:ncol(data_merge$sample.info))], 1, function(x) mean(x, na.rm = TRUE))

data.plot <- 
  cbind(data_all, Sample.mean) %>% 
  mutate(ID = row_number(),
         Sample.mean = scale(Sample.mean),
         Batch = ifelse(Sample.type %in% params$po.sample.to.use, "QC", Batch))

p <- 
  ggplot(data.plot, aes(x = ID, y = Sample.mean, color = Batch, text = Sample)) +
    geom_point() +
    theme_bw()  +
    scale_color_npg()

ggplotly(p)

```

$~$

# 4 Metabolites summary {.tabset .tabset-pills}
## Metabolites class
```{r, fig.height = 4, fig.width = 8}
features <- 
  data_merge$feature.info

if (!"Metabolite.class" %in% names(features)) features$Metabolite.class <- features$main_class

data.plot <- data.frame(table(features$Metabolite.class))
names(data.plot)[1] <- "Class"

plot_ly(data.plot, labels = ~Class, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```

## Identification level
```{r, fig.height = 4, fig.width = 8}
data.plot <- 
  data.frame(table(features$Source)) %>% 
  mutate(Source = case_when( Var1 == "in_house_MS1_matched" ~ "Level 1",
                             Var1 == "public_db_MS2_matched" ~ "Level 2",
                             Var1 == "public_db_mz_matched" ~ "m/z match",
                             TRUE ~ "Others"))


plot_ly(data.plot, labels = ~Source, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```




