---
title: "Data preprocessing evaluation"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '3'
    number_sections: true
    df_print: paged

params:
  config: ""
---
<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE, warning = FALSE}

library(dplyr)
library(tibble)
library(plotly)
library(ggsci)
library(yaml)

load_metpipe_runtime <- function() {
  package_mode <- tolower(trimws(Sys.getenv("METPIPE_PACKAGE_MODE", unset = "local")))
  if (!package_mode %in% c("local", "installed")) package_mode <- "local"

  if (identical(package_mode, "installed")) {
    if (!requireNamespace("metpipe", quietly = TRUE)) {
      stop("METPIPE_PACKAGE_MODE='installed' but package 'metpipe' is not installed")
    }
    library(metpipe)
    return("installed")
  }

  candidates <- unique(Filter(nzchar, c(
    Sys.getenv("METPIPE_LOCAL_ROOT", ""),
    "H:/Documents/CBMR_workflow/packages/metpipe",
    normalizePath(getwd(), winslash = "/", mustWork = FALSE),
    normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), winslash = "/", mustWork = FALSE)
  )))

  is_pkg_root <- function(path) {
    dir.exists(path) &&
      file.exists(file.path(path, "DESCRIPTION")) &&
      dir.exists(file.path(path, "R"))
  }

  roots <- candidates[vapply(candidates, is_pkg_root, logical(1))]
  if (length(roots) == 0) {
    stop("Cannot locate local metpipe package root. Set METPIPE_LOCAL_ROOT to your metpipe folder.")
  }

  if (!requireNamespace("pkgload", quietly = TRUE)) {
    stop("pkgload is required for local source loading. Install with install.packages('pkgload').")
  }

  Sys.setenv(METPIPE_LOCAL_ROOT = roots[[1]])
  pkgload::load_all(path = roots[[1]], export_all = FALSE, quiet = TRUE)
  "local"
}

metpipe_runtime <- load_metpipe_runtime()

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

is_blank <- function(x) {
  is.null(x) || length(x) == 0 || all(is.na(x)) || !nzchar(trimws(as.character(x[1])))
}

has_cols <- function(df, cols) {
  !is.null(df) && is.data.frame(df) && all(cols %in% names(df))
}
```

```{r pipeline input}

if (is_blank(params$config)) {
  stop("params$config is required")
}
if (!file.exists(params$config)) {
  stop("Config file does not exist: ", params$config)
}

config <- read_yaml(params$config)
if (is.null(config) || !is.list(config)) {
  stop("Invalid YAML config: expected a named list")
}

rds_datalist <- config$rds_datalist
if (is_blank(rds_datalist)) {
  stop("Missing required config field: rds_datalist")
}
if (!file.exists(rds_datalist)) {
  stop("datalist file not found: ", rds_datalist)
}

```


```{r}

datalist <- readRDS(rds_datalist)

if (!is.list(datalist)) {
  stop("Invalid datalist object: expected a list")
}

pos <- datalist$pos
neg <- datalist$neg

run.clean.pos <- !is.null(pos) && !is.null(pos$peaks_cleaned) && has_cols(pos$features_cleaned, c("Identity", "Feature_type"))
run.clean.neg <- !is.null(neg) && !is.null(neg$peaks_cleaned) && has_cols(neg$features_cleaned, c("Identity", "Feature_type"))

run.istd.pos <- run.clean.pos && any(pos$features_cleaned$Feature_type == "IS", na.rm = TRUE)
run.istd.neg <- run.clean.neg && any(neg$features_cleaned$Feature_type == "IS", na.rm = TRUE)

run.norm.pos <- !is.null(pos) && !is.null(pos$peaks_norm) && length(pos$peaks_norm) > 0 && has_cols(pos$meta_norm, c("Sample", "Sample.type", "Sample.batch"))
run.norm.neg <- !is.null(neg) && !is.null(neg$peaks_norm) && length(neg$peaks_norm) > 0 && has_cols(neg$meta_norm, c("Sample", "Sample.type", "Sample.batch"))

run.merged <- !is.null(datalist$datatable) && has_cols(datalist$sample.info, c("Sample", "Sample.batch", "Run.order"))
run.feature.summary <- !is.null(datalist$feature.info) && is.data.frame(datalist$feature.info)

```

# Positive mode
## Missing distribution {.tabset .tabset-pills}
```{r, eval = run.clean.pos}

data_cleaned <- datalist$pos$peaks_cleaned
rownames(data_cleaned) <- datalist$pos$features_cleaned$Identity

figs <- 
  missing_eval(data_cleaned = data_cleaned)
```

### Overall
```{r, eval = run.clean.pos, fig.width = 11, fig.height = 5}
figs$p1
```

### Missing in samples
```{r, eval = run.clean.pos, fig.width = 9, fig.height = 4}
ggplotly(figs$p2)
```

### Missing in features
```{r, eval = run.clean.pos, fig.width = 9, fig.height = 4}
ggplotly(figs$p3)
```

## Internal standards {.tabset}
```{r eval = run.istd.pos}

istd.index <- datalist$pos$features_cleaned$Feature_type == "IS"
features.istd <- datalist$pos$features_cleaned[istd.index, ]
istd <- datalist$pos$peaks_cleaned[istd.index, ]

data.istd <- data.frame(t(istd))
names(data.istd) <- features.istd$Identity_raw

data.input <- 
  list(data.istd = data.istd, 
       sample.info = datalist$pos$meta_cleaned)

plt.is.rsd <-
  plot_istd_rsd(datalist = data.input)

```

### RSD
```{r rsd.bar.plot pos, fig.height = 6, fig.width = 8, eval = run.istd.pos}

ggplotly(plt.is.rsd$plt.rsd)

```

### IS intensities in samples
```{r rsd.sample pos, eval = run.istd.pos}

plt.is.rsd$plt.int

```

## Normalization comparison
```{r, eval = run.norm.pos}

pca.plot <- 
  plot_list_pca(
    x = datalist$pos$peaks_norm,
    group = datalist$pos$meta_norm$Sample.batch,
    group.label = "Batch",
    plotly.text = datalist$pos$meta_norm$Sample)

pca.plot.type <- 
  plot_list_pca(
    x = datalist$pos$peaks_norm,
    group = datalist$pos$meta_norm$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$pos$meta_norm$Sample)

plt.eval <-
  plot_eval( x = datalist$pos$peaks_norm,
             type = datalist$pos$meta_norm$Sample.type)

```

### PCA {.tabset .tabset-pills}

#### Raw Data

```{r PCA-Loaded pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

a <- ggplotly(pca.plot$raw, session="knitr", showlegend = T)
b <- ggplotly(pca.plot.type$raw, session="knitr", showlegend = T)

subplot(a, b)
```

#### Best IS

```{r PCA-Bestid pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$bestis)){

  a <- ggplotly(pca.plot$bestis, session="knitr")
  b <- ggplotly(pca.plot.type$bestis, session="knitr")
  
  subplot(a, b)
}

```

#### Lowest CV

```{r PCA-Lowest CV pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$low_cv)){

  a <- ggplotly(pca.plot$low_cv, session="knitr")
  b <- ggplotly(pca.plot.type$low_cv, session="knitr")
  
  subplot(a, b)
}

```

#### Sum

```{r PCA-Normalized_by_sum pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$sum)){

  a <- ggplotly(pca.plot$sum, session="knitr")
  b <- ggplotly(pca.plot.type$sum, session="knitr")
  
  subplot(a, b)
}
```

#### Median
```{r PCA-Normalized_by_median pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$median)){

  a <- ggplotly(pca.plot$median, session="knitr")
  b <- ggplotly(pca.plot.type$median, session="knitr")
  
  subplot(a, b)
}
```

#### PQN
```{r PCA-Normalized_by_PQN pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$pqn)){

  a <- ggplotly(pca.plot$pqn, session="knitr")
  b <- ggplotly(pca.plot.type$pqn, session="knitr")
  
  subplot(a, b)
}
```

#### Loess
```{r PCA-Normalized_by_loess pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$loess)){
  a <- ggplotly(pca.plot$loess, session="knitr")
  b <- ggplotly(pca.plot.type$loess, session="knitr")
  
  subplot(a, b)
}
```

#### Combat
```{r PCA-Normalized_by_combat pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$combat)){

  a <- ggplotly(pca.plot$combat, session="knitr")
  b <- ggplotly(pca.plot.type$combat, session="knitr")
  
  subplot(a, b)
}

```

#### limma
```{r PCA-Normalized_by_limma pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

if (!is.null(pca.plot$limma)){

  a <- ggplotly(pca.plot$limma, session="knitr")
  b <- ggplotly(pca.plot.type$limma, session="knitr")
  
  subplot(a, b)
}

```

### RSD for all features in QC samples {.tabset .tabset-pills}

#### RSD
```{r RSD pos, fig.height = 3, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.rsd
plt.eval$plt.rsd.mean

```

#### MAD
```{r MAD pos, fig.height = 3, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.mad
plt.eval$plt.mad.mean

```

#### Variance
```{r VAR pos, fig.height = 3, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.var
plt.eval$plt.var.sum

```
$~$

# Negative mode
## Missing distribution {.tabset .tabset-pills}
```{r, eval = run.clean.neg}

data_cleaned <- datalist$neg$peaks_cleaned
rownames(data_cleaned) <- datalist$neg$features_cleaned$Identity

figs <- 
  missing_eval(data_cleaned = data_cleaned)
```

### Overall
```{r, eval = run.clean.neg, fig.width = 11, fig.height = 5}
figs$p1
```

### Missing in samples
```{r, eval = run.clean.neg, fig.width = 9, fig.height = 4}
ggplotly(figs$p2)
```

### Missing in features
```{r, eval = run.clean.neg, fig.width = 9, fig.height = 4}
ggplotly(figs$p3)
```

## Internal standards {.tabset}
```{r eval = run.istd.neg}

istd.index <- datalist$neg$features_cleaned$Feature_type == "IS"
features.istd <- datalist$neg$features_cleaned[istd.index, ]
istd <- datalist$neg$peaks_cleaned[istd.index, ]

data.istd <- data.frame(t(istd))
names(data.istd) <- features.istd$Identity_raw

data.input <- 
  list(data.istd = data.istd, 
       sample.info = datalist$neg$meta_cleaned)

plt.is.rsd <-
  plot_istd_rsd(datalist = data.input)

```

### RSD
```{r rsd.bar.plot neg, fig.height = 6, fig.width = 8, eval = run.istd.neg}

ggplotly(plt.is.rsd$plt.rsd)

```

### IS intensities in samples
```{r rsd.sample neg, eval = run.istd.neg}

plt.is.rsd$plt.int

```

## Normalization comparison
```{r, eval = run.norm.neg}

pca.plot <- 
  plot_list_pca(
    x = datalist$neg$peaks_norm,
    group = datalist$neg$meta_norm$Sample.batch,
    group.label = "Batch",
    plotly.text = datalist$neg$meta_norm$Sample)

pca.plot.type <- 
  plot_list_pca(
    x = datalist$neg$peaks_norm,
    group = datalist$neg$meta_norm$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$neg$meta_norm$Sample)

plt.eval <-
  plot_eval( x = datalist$neg$peaks_norm,
             type = datalist$neg$meta_norm$Sample.type)

```

### PCA {.tabset .tabset-pills}

#### Raw Data

```{r PCA-Loaded neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

a <- ggplotly(pca.plot$raw, session="knitr", showlegend = T)
b <- ggplotly(pca.plot.type$raw, session="knitr", showlegend = T)

subplot(a, b)
```

#### Best IS

```{r PCA-Bestid neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$bestis)){

  a <- ggplotly(pca.plot$bestis, session="knitr")
  b <- ggplotly(pca.plot.type$bestis, session="knitr")
  
  subplot(a, b)
}

```

#### Lowest CV

```{r PCA-Lowest CV neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$low_cv)){

  a <- ggplotly(pca.plot$low_cv, session="knitr")
  b <- ggplotly(pca.plot.type$low_cv, session="knitr")
  
  subplot(a, b)
}

```

#### Sum

```{r PCA-Normalized_by_sum neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$sum)){

  a <- ggplotly(pca.plot$sum, session="knitr")
  b <- ggplotly(pca.plot.type$sum, session="knitr")
  
  subplot(a, b)
}
```

#### Median
```{r PCA-Normalized_by_median neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$median)){

  a <- ggplotly(pca.plot$median, session="knitr")
  b <- ggplotly(pca.plot.type$median, session="knitr")
  
  subplot(a, b)
}
```

#### PQN
```{r PCA-Normalized_by_PQN neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$pqn)){

  a <- ggplotly(pca.plot$pqn, session="knitr")
  b <- ggplotly(pca.plot.type$pqn, session="knitr")
  
  subplot(a, b)
}
```

#### Loess
```{r PCA-Normalized_by_loess neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$loess)){
  a <- ggplotly(pca.plot$loess, session="knitr")
  b <- ggplotly(pca.plot.type$loess, session="knitr")
  
  subplot(a, b)
}
```

#### Combat
```{r PCA-Normalized_by_combat neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$combat)){

  a <- ggplotly(pca.plot$combat, session="knitr")
  b <- ggplotly(pca.plot.type$combat, session="knitr")
  
  subplot(a, b)
}

```

#### limma
```{r PCA-Normalized_by_limma neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

if (!is.null(pca.plot$limma)){

  a <- ggplotly(pca.plot$limma, session="knitr")
  b <- ggplotly(pca.plot.type$limma, session="knitr")
  
  subplot(a, b)
}

```

### RSD for all features in QC samples {.tabset .tabset-pills}

#### RSD
```{r RSD neg, fig.height = 3, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.rsd
plt.eval$plt.rsd.mean

```

#### MAD
```{r MAD neg, fig.height = 3, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.mad
plt.eval$plt.mad.mean

```

#### Variance
```{r VAR neg, fig.height = 3, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.var
plt.eval$plt.var.sum

```
$~$

# Merged data {.tabset .tabset-pills}
## PCA
```{r, fig.height = 4, fig.width = 6, eval = run.merged}

data_all <- datalist$datatable
sample.info <- datalist$sample.info

p <- 
  plot_PCA(
      data_all[, -c(1:ncol(sample.info))],
      group = sample.info$Sample.batch,
      group.label = "Batch",
      plotly.text = sample.info$Sample,
      ratio = 1)
  
ggplotly(p)

```

## Sample mean across batches
```{r, fig.height = 4, fig.width = 8, eval = run.merged}

Sample.mean <- 
  apply(data_all[, -c(1:ncol(sample.info))], 1, function(x) mean(x, na.rm = TRUE))

# data.plot <- 
#   cbind(data_all, Sample.mean) %>% 
#   mutate(ID = row_number(),
#          Sample.mean = scale(Sample.mean),
#          Batch = ifelse(Sample.type %in% params$po.sample.to.use, "QC", Batch))

scaled <- 
  apply(data_all[, -c(1:ncol(sample.info))], 2, function(x) scale(x)) %>% 
  as.data.frame()

data.plot <- 
  sample.info %>% 
  cbind(., scaled) %>% 
  pivot_longer(-c(1:(ncol(sample.info) + 1)), names_to = "met", values_to = "value")

p <- 
  ggplot(data.plot, aes(x = Run.order, y = value, color = Sample.batch, text = Sample)) +
    geom_boxplot() +
    theme_bw()  +
    ggsci::scale_color_npg()

ggplotly(p)

```


# Metabolites summary {.tabset .tabset-pills}
## Metabolites class
```{r, fig.height = 4, fig.width = 8, eval = run.feature.summary}
features <- 
  datalist$feature.info

if (!"Metabolite.class" %in% names(features)) features$Metabolite.class <- features$main_class

data.plot <- data.frame(table(features$Metabolite.class))
names(data.plot)[1] <- "Class"

if (nrow(data.plot) == 0) data.plot <- data.frame(Class = NA, Freq = nrow(features))

plot_ly(data.plot, labels = ~Class, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```

## Identification level
```{r, fig.height = 4, fig.width = 8, eval = run.feature.summary}
data.plot <- 
  data.frame(table(features$identity_source, useNA = "ifany")) %>% 
  mutate(Source = case_when( !is.na(Var1) ~ Var1,
                             TRUE ~ "Unknown"))


plot_ly(data.plot, labels = ~Source, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```

```{r session info, include = FALSE}

# Session info

utils::sessionInfo()

```