---
title: "Data preprocessing evaluation"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
          
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    theme: flatly
    polyfill: false
    page-layout: full
    number-sections: true
    number-depth: 2
self-contained: true

params:
  config: ""
---

```{r setup environment, include = FALSE, warning = FALSE}

library(dplyr)
library(tidyr)
library(tibble)
library(plotly)
library(ggsci)
library(yaml)

helper_candidates <- c(
  file.path(getwd(), "helpers_workflow.R"),
  file.path(getwd(), "inst", "scripts", "helpers_workflow.R"),
  system.file("scripts", "helpers_workflow.R", package = "metpipe")
)
helper_candidates <- helper_candidates[file.exists(helper_candidates)]
if (length(helper_candidates) == 0) {
  stop("Cannot locate shared helper script helpers_workflow.R")
}
source(helper_candidates[[1]], local = TRUE)

metpipe_runtime <- wf_load_metpipe_runtime()

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

is_blank <- wf_is_blank

has_cols <- function(df, cols) {
  !is.null(df) && is.data.frame(df) && all(cols %in% names(df))
}
```

```{r pipeline input}

config <- wf_read_yaml_config(params$config, label = "params$config")

rds_datalist <- wf_cfg_required(config, "rds_datalist")
if (!file.exists(rds_datalist)) {
  stop("datalist file not found: ", rds_datalist)
}

norm_methods_selected <- config$norm.method
if (is.null(norm_methods_selected)) {
  norm_methods_selected <- character(0)
}
if (length(norm_methods_selected) == 1 && grepl(",", norm_methods_selected, fixed = TRUE)) {
  norm_methods_selected <- trimws(unlist(strsplit(norm_methods_selected, ",", fixed = TRUE)))
}
norm_methods_selected <- unique(tolower(as.character(norm_methods_selected[!is.na(norm_methods_selected)])))

is_norm_selected <- function(method_name) {
  tolower(method_name) %in% norm_methods_selected
}

show.bestis <- is_norm_selected("bestis")
show.low_cv <- is_norm_selected("low_cv")
show.sum <- is_norm_selected("sum")
show.median <- is_norm_selected("median")
show.pqn <- is_norm_selected("pqn")
show.loess <- is_norm_selected("loess")
show.combat <- is_norm_selected("combat")
show.limma <- is_norm_selected("limma")

```


```{r}

datalist <- readRDS(rds_datalist)

if (!is.list(datalist)) {
  stop("Invalid datalist object: expected a list")
}

datalist <- as_metpipe_datalist(datalist, validate = FALSE)
pos <- get_mode(datalist, "pos", required = FALSE)
neg <- get_mode(datalist, "neg", required = FALSE)

run.clean.pos <- !is.null(pos) && !is.null(pos$peaks_cleaned) && has_cols(pos$features_cleaned, c("Identity", "Feature_type"))
run.clean.neg <- !is.null(neg) && !is.null(neg$peaks_cleaned) && has_cols(neg$features_cleaned, c("Identity", "Feature_type"))

run.istd.pos <- run.clean.pos && any(pos$features_cleaned$Feature_type == "IS", na.rm = TRUE)
run.istd.neg <- run.clean.neg && any(neg$features_cleaned$Feature_type == "IS", na.rm = TRUE)

run.norm.pos <- !is.null(pos) && !is.null(pos$peaks_norm) && length(pos$peaks_norm) > 0 && has_cols(pos$meta_norm, c("Sample", "Sample.type", "Sample.batch"))
run.norm.neg <- !is.null(neg) && !is.null(neg$peaks_norm) && length(neg$peaks_norm) > 0 && has_cols(neg$meta_norm, c("Sample", "Sample.type", "Sample.batch"))

run.merged <- !is.null(datalist$datatable) && has_cols(datalist$sample.info, c("Sample", "Sample.batch", "Run.order"))
run.feature.summary <- !is.null(datalist$feature.info) && is.data.frame(datalist$feature.info)

```

## Positive mode
### Missing distribution
::: panel-tabset
```{r, eval = run.clean.pos}

data_cleaned <- datalist$pos$peaks_cleaned
rownames(data_cleaned) <- datalist$pos$features_cleaned$Identity

figs <- 
  missing_eval(data_cleaned = data_cleaned)
```

#### Overall
```{r, eval = run.clean.pos, fig.height = 5}
figs$p1

```

#### Missing in samples
```{r, eval = run.clean.pos, fig.height = 5}
ggplotly(figs$p2) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

#### Missing in features
```{r, eval = run.clean.pos, fig.height = 5}
ggplotly(figs$p3) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )
```
:::

### Internal standards
::: panel-tabset
```{r eval = run.istd.pos}

istd.index <- datalist$pos$features_cleaned$Feature_type == "IS"
features.istd <- datalist$pos$features_cleaned[istd.index, ]
istd <- datalist$pos$peaks_cleaned[istd.index, ]

data.istd <- data.frame(t(istd))
names(data.istd) <- features.istd$Identity_raw

data.input <- 
  list(data.istd = data.istd, 
       sample.info = datalist$pos$meta_cleaned)

plt.is.rsd <-
  plot_istd_rsd(datalist = data.input)

```

#### RSD
```{r rsd.bar.plot pos, fig.height = 6, fig.width = 8, eval = run.istd.pos, out.width = "80%"}

ggplotly(plt.is.rsd$plt.rsd) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```

#### IS intensities in samples
```{r rsd.sample pos, eval = run.istd.pos}

plt.is.rsd$plt.int %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```
:::

### Normalization comparison

```{r, eval = run.norm.pos}

pca.plot <- 
  plot_list_pca(
    x = datalist$pos$peaks_norm,
    group = datalist$pos$meta_norm$Sample.batch,
    group.label = "Batch",
    plotly.text = datalist$pos$meta_norm$Sample)

pca.plot.type <- 
  plot_list_pca(
    x = datalist$pos$peaks_norm,
    group = datalist$pos$meta_norm$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$pos$meta_norm$Sample)

plt.eval <-
  plot_eval( x = datalist$pos$peaks_norm,
             type = datalist$pos$meta_norm$Sample.type)

```

#### PCA
::: panel-tabset

##### Raw Data

```{r PCA-Loaded pos, fig.height = 5, fig.width = 8, eval = run.norm.pos}

a <- ggplotly(pca.plot$raw, session="knitr", showlegend = T) 
b <- ggplotly(pca.plot.type$raw, session="knitr", showlegend = T) 

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-bestis-pos, results='asis', eval = run.norm.pos && show.bestis && !is.null(pca.plot$bestis)}
cat("##### Best IS\n")
```

```{r PCA-Bestid pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.bestis && !is.null(pca.plot$bestis)}

a <- ggplotly(pca.plot$bestis, session="knitr")
b <- ggplotly(pca.plot.type$bestis, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-lowcv-pos, results='asis', eval = run.norm.pos && show.low_cv && !is.null(pca.plot$low_cv)}
cat("##### Lowest CV\n")
```

```{r PCA-Lowest CV pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.low_cv && !is.null(pca.plot$low_cv)}

a <- ggplotly(pca.plot$low_cv, session="knitr")
b <- ggplotly(pca.plot.type$low_cv, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```

```{r pca-title-sum-pos, results='asis', eval = run.norm.pos && show.sum && !is.null(pca.plot$sum)}
cat("##### Sum\n")
```

```{r PCA-Normalized_by_sum pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.sum && !is.null(pca.plot$sum)}

a <- ggplotly(pca.plot$sum, session="knitr")
b <- ggplotly(pca.plot.type$sum, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-median-pos, results='asis', eval = run.norm.pos && show.median && !is.null(pca.plot$median)}
cat("##### Median\n")
```

```{r PCA-Normalized_by_median pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.median && !is.null(pca.plot$median)}

a <- ggplotly(pca.plot$median, session="knitr")
b <- ggplotly(pca.plot.type$median, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-pqn-pos, results='asis', eval = run.norm.pos && show.pqn && !is.null(pca.plot$pqn)}
cat("##### PQN\n")
```

```{r PCA-Normalized_by_PQN pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.pqn && !is.null(pca.plot$pqn)}

a <- ggplotly(pca.plot$pqn, session="knitr")
b <- ggplotly(pca.plot.type$pqn, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-loess-pos, results='asis', eval = run.norm.pos && show.loess && !is.null(pca.plot$loess)}
cat("##### Loess\n")
```

```{r PCA-Normalized_by_loess pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.loess && !is.null(pca.plot$loess)}

a <- ggplotly(pca.plot$loess, session="knitr")
b <- ggplotly(pca.plot.type$loess, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-combat-pos, results='asis', eval = run.norm.pos && show.combat && !is.null(pca.plot$combat)}
cat("##### Combat\n")
```

```{r PCA-Normalized_by_combat pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.combat && !is.null(pca.plot$combat)}

a <- ggplotly(pca.plot$combat, session="knitr")
b <- ggplotly(pca.plot.type$combat, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-limma-pos, results='asis', eval = run.norm.pos && show.limma && !is.null(pca.plot$limma)}
cat("##### limma\n")
```

```{r PCA-Normalized_by_limma pos, fig.height = 5, fig.width = 8, eval = run.norm.pos && show.limma && !is.null(pca.plot$limma)}

a <- ggplotly(pca.plot$limma, session="knitr")
b <- ggplotly(pca.plot.type$limma, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```
:::

#### RSD for all features in QC samples
::: panel-tabset

##### RSD
```{r RSD pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.rsd
plt.eval$plt.rsd.mean

```

##### MAD
```{r MAD pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.mad
plt.eval$plt.mad.mean

```

##### Variance
```{r VAR pos, fig.height = 4, fig.width = 8, eval = run.norm.pos}

plt.eval$plt.var
plt.eval$plt.var.sum

```
$~$
:::

## Negative mode
### Missing distribution
::: panel-tabset
```{r, eval = run.clean.neg}

data_cleaned <- datalist$neg$peaks_cleaned
rownames(data_cleaned) <- datalist$neg$features_cleaned$Identity

figs <- 
  missing_eval(data_cleaned = data_cleaned)
```

#### Overall
```{r, eval = run.clean.neg, fig.height = 5}
figs$p1
```

#### Missing in samples
```{r, eval = run.clean.neg, fig.height = 5}
ggplotly(figs$p2) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )
```

#### Missing in features
```{r, eval = run.clean.neg, fig.height = 5}
ggplotly(figs$p3) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )
```
:::

### Internal standards
::: panel-tabset

```{r eval = run.istd.neg}

istd.index <- datalist$neg$features_cleaned$Feature_type == "IS"
features.istd <- datalist$neg$features_cleaned[istd.index, ]
istd <- datalist$neg$peaks_cleaned[istd.index, ]

data.istd <- data.frame(t(istd))
names(data.istd) <- features.istd$Identity_raw

data.input <- 
  list(data.istd = data.istd, 
       sample.info = datalist$neg$meta_cleaned)

plt.is.rsd <-
  plot_istd_rsd(datalist = data.input)

```

#### RSD
```{r rsd.bar.plot neg, fig.height = 6, fig.width = 8, eval = run.istd.neg, out.width = "80%"}

ggplotly(plt.is.rsd$plt.rsd) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

#### IS intensities in samples
```{r rsd.sample neg, eval = run.istd.neg}

plt.is.rsd$plt.int %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```
:::

### Normalization comparison
```{r, eval = run.norm.neg}

pca.plot <- 
  plot_list_pca(
    x = datalist$neg$peaks_norm,
    group = datalist$neg$meta_norm$Sample.batch,
    group.label = "Batch",
    plotly.text = datalist$neg$meta_norm$Sample)

pca.plot.type <- 
  plot_list_pca(
    x = datalist$neg$peaks_norm,
    group = datalist$neg$meta_norm$Sample.type,
    group.label = "Sample type",
    plotly.text = datalist$neg$meta_norm$Sample)

plt.eval <-
  plot_eval( x = datalist$neg$peaks_norm,
             type = datalist$neg$meta_norm$Sample.type)

```

#### PCA
::: panel-tabset

##### Raw Data

```{r PCA-Loaded neg, fig.height = 5, fig.width = 8, eval = run.norm.neg}

a <- ggplotly(pca.plot$raw, session="knitr", showlegend = T)
b <- ggplotly(pca.plot.type$raw, session="knitr", showlegend = T)

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )
```

```{r pca-title-bestis-neg, results='asis', eval = run.norm.neg && show.bestis && !is.null(pca.plot$bestis)}
cat("##### Best IS\n")
```

```{r PCA-Bestid neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.bestis && !is.null(pca.plot$bestis)}

a <- ggplotly(pca.plot$bestis, session="knitr")
b <- ggplotly(pca.plot.type$bestis, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```

```{r pca-title-lowcv-neg, results='asis', eval = run.norm.neg && show.low_cv && !is.null(pca.plot$low_cv)}
cat("##### Lowest CV\n")
```

```{r PCA-Lowest CV neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.low_cv && !is.null(pca.plot$low_cv)}

a <- ggplotly(pca.plot$low_cv, session="knitr")
b <- ggplotly(pca.plot.type$low_cv, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```

```{r pca-title-sum-neg, results='asis', eval = run.norm.neg && show.sum && !is.null(pca.plot$sum)}
cat("##### Sum\n")
```

```{r PCA-Normalized_by_sum neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.sum && !is.null(pca.plot$sum)}

a <- ggplotly(pca.plot$sum, session="knitr")
b <- ggplotly(pca.plot.type$sum, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-median-neg, results='asis', eval = run.norm.neg && show.median && !is.null(pca.plot$median)}
cat("##### Median\n")
```

```{r PCA-Normalized_by_median neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.median && !is.null(pca.plot$median)}

a <- ggplotly(pca.plot$median, session="knitr")
b <- ggplotly(pca.plot.type$median, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-pqn-neg, results='asis', eval = run.norm.neg && show.pqn && !is.null(pca.plot$pqn)}
cat("##### PQN\n")
```

```{r PCA-Normalized_by_PQN neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.pqn && !is.null(pca.plot$pqn)}

a <- ggplotly(pca.plot$pqn, session="knitr")
b <- ggplotly(pca.plot.type$pqn, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-loess-neg, results='asis', eval = run.norm.neg && show.loess && !is.null(pca.plot$loess)}
cat("##### Loess\n")
```

```{r PCA-Normalized_by_loess neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.loess && !is.null(pca.plot$loess)}

a <- ggplotly(pca.plot$loess, session="knitr")
b <- ggplotly(pca.plot.type$loess, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

```{r pca-title-combat-neg, results='asis', eval = run.norm.neg && show.combat && !is.null(pca.plot$combat)}
cat("##### Combat\n")
```

```{r PCA-Normalized_by_combat neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.combat && !is.null(pca.plot$combat)}

a <- ggplotly(pca.plot$combat, session="knitr")
b <- ggplotly(pca.plot.type$combat, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```

```{r pca-title-limma-neg, results='asis', eval = run.norm.neg && show.limma && !is.null(pca.plot$limma)}
cat("##### limma\n")
```

```{r PCA-Normalized_by_limma neg, fig.height = 5, fig.width = 8, eval = run.norm.neg && show.limma && !is.null(pca.plot$limma)}

a <- ggplotly(pca.plot$limma, session="knitr")
b <- ggplotly(pca.plot.type$limma, session="knitr")

subplot(a, b) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    xaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis2 = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )


```
:::

#### RSD for all features in QC samples
::: panel-tabset

##### RSD
```{r RSD neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.rsd
plt.eval$plt.rsd.mean

```

##### MAD
```{r MAD neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.mad
plt.eval$plt.mad.mean

```

##### Variance
```{r VAR neg, fig.height = 4, fig.width = 8, eval = run.norm.neg}

plt.eval$plt.var
plt.eval$plt.var.sum

```
$~$
:::

## Merged data
::: panel-tabset

### PCA
```{r, fig.height = 5, fig.width = 6, eval = run.merged, out.width = "80%"}

data_all <- datalist$datatable
sample.info <- datalist$sample.info

p <- 
  plot_PCA(
      data_all[, -c(1:ncol(sample.info))],
      group = sample.info$Sample.batch,
      group.label = "Batch",
      plotly.text = sample.info$Sample,
      ratio = 1)
  
ggplotly(p) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```

### Sample mean across batches
```{r, fig.height = 5, fig.width = 8, eval = run.merged}

Sample.mean <- 
  apply(data_all[, -c(1:ncol(sample.info))], 1, function(x) mean(x, na.rm = TRUE))

# data.plot <- 
#   cbind(data_all, Sample.mean) %>% 
#   mutate(ID = row_number(),
#          Sample.mean = scale(Sample.mean),
#          Batch = ifelse(Sample.type %in% params$po.sample.to.use, "QC", Batch))

scaled <- 
  apply(data_all[, -c(1:ncol(sample.info))], 2, function(x) scale(x)) %>% 
  as.data.frame()

data.plot <- 
  sample.info %>% 
  cbind(., scaled) %>% 
  pivot_longer(-c(1:(ncol(sample.info) + 1)), names_to = "met", values_to = "value")

p <- 
  ggplot(data.plot, aes(x = Run.order, y = value, color = Sample.batch, text = Sample)) +
    geom_boxplot() +
    theme_bw()  +
    ggsci::scale_color_npg()

ggplotly(p) %>%
  layout(
    xaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all"),
    yaxis = list(showline = TRUE, linewidth = 1, linecolor = '#999999', mirror = "all")
  )

```
:::

## Metabolites summary
::: panel-tabset

### Metabolites class
```{r, fig.height = 4, fig.width = 8, eval = run.feature.summary, out.width = "80%"}
features <- 
  datalist$feature.info

if (!"Metabolite.class" %in% names(features)) features$Metabolite.class <- features$main_class

data.plot <- data.frame(table(features$Metabolite.class))
names(data.plot)[1] <- "Class"

if (nrow(data.plot) == 0) data.plot <- data.frame(Class = NA, Freq = nrow(features))

plot_ly(data.plot, labels = ~Class, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```

### Identification level
```{r, fig.height = 4, fig.width = 8, eval = run.feature.summary, out.width = "80%"}
data.plot <- 
  data.frame(table(features$identity_source, useNA = "ifany")) %>% 
  mutate(Source = case_when( !is.na(Var1) ~ Var1,
                             TRUE ~ "Unknown"))


plot_ly(data.plot, labels = ~Source, values = ~Freq, type = 'pie',
               marker = list(line = list(color = '#FFFFFF', width = 1)))

```
:::

```{r session info, include = FALSE, eval = FALSE}

# Session info

utils::sessionInfo()

```