---
title: "Merge batch evaluation"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    theme: flatly
    polyfill: false
    page-layout: full
    number-sections: true
    number-depth: 2
self-contained: true

params:
  config: ""
  n_batches: 2
  overlap_by_name: ""
---

```{r setup, include=FALSE, warning=FALSE}

library(yaml)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(openxlsx)

helper_candidates <- c(
  file.path(getwd(), "helpers_workflow.R"),
  file.path(getwd(), "inst", "scripts", "helpers_workflow.R"),
  system.file("scripts", "helpers_workflow.R", package = "metpipe")
)
helper_candidates <- helper_candidates[file.exists(helper_candidates)]
if (length(helper_candidates) == 0) stop("Cannot locate helpers_workflow.R")
source(helper_candidates[[1]], local = TRUE)

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, include = TRUE)

```

```{r functions, include=FALSE}

# ── Feature subset helpers ────────────────────────────────────────────────────

make_feature_subsets <- function(diag_df) {
  list(
    `Internal standard` = diag_df[grepl("istd", diag_df$table), , drop = FALSE],
    `With identity`     = diag_df[!grepl("istd", diag_df$table) & !is.na(diag_df$identity), , drop = FALSE],
    `Without identity`  = diag_df[!grepl("istd", diag_df$table) &  is.na(diag_df$identity), , drop = FALSE],
    `Overall`           = diag_df
  )
}

make_l2r_subsets <- function(l2r_df) {
  empty <- data.frame(table = character(), step = character(),
                      identity = character(), sample = character(),
                      log2ratio = numeric(), stringsAsFactors = FALSE)
  if (is.null(l2r_df) || nrow(l2r_df) == 0)
    return(list(`Internal standard` = empty, `With identity` = empty,
                `Without identity`  = empty, `Overall`       = empty))
  is_istd   <- grepl("istd", l2r_df$table)
  has_id_col <- "identity" %in% colnames(l2r_df)
  list(
    `Internal standard` = l2r_df[ is_istd, , drop = FALSE],
    `With identity`     = if (has_id_col)
      l2r_df[!is_istd & !is.na(l2r_df$identity), , drop = FALSE]
    else
      l2r_df[!is_istd, , drop = FALSE],
    `Without identity`  = if (has_id_col)
      l2r_df[!is_istd &  is.na(l2r_df$identity), , drop = FALSE]
    else
      empty,
    `Overall`           = l2r_df
  )
}

# ── Alignment quality plots ───────────────────────────────────────────────────

plot_mz_hist <- function(df) {
  if (nrow(df) == 0) { cat("No data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(df, aes(x = mz_ppm)) +
    geom_histogram(bins = 60, fill = "steelblue", color = "white", linewidth = 0.2) +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(#title = "m/z difference for matched feature pairs",
         x = "m/z difference (ppm)", y = "Nr. features")
  ggplotly(p)
}

plot_rt_hist <- function(df) {
  if (nrow(df) == 0) { cat("No data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(df, aes(x = rt_diff)) +
    geom_histogram(bins = 60, fill = "darkorange", color = "white", linewidth = 0.2) +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(#title = "RT difference for matched feature pairs",
         x = "|RT difference| (min)", y = "Nr. features")
  ggplotly(p)
}

# ── Overlap consistency plots ─────────────────────────────────────────────────

plot_cor_hist <- function(df) {
  d <- df[!is.na(df$overlap_cor), , drop = FALSE]
  if (nrow(d) == 0) { cat("No overlap correlation data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(d, aes(x = overlap_cor)) +
    geom_histogram(bins = 60, fill = "seagreen", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = 0.8, linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(#title = "Overlap sample Spearman correlation per feature (batch A vs batch B)",
         subtitle = "Red dashed line: 0.8 quality threshold",
         x = "Spearman r", y = "Nr. features")
  ggplotly(p)
}

print_cor_summary <- function(df) {
  d <- df[!is.na(df$overlap_cor), , drop = FALSE]
  if (nrow(d) == 0) return(invisible(NULL))
  summ <- d |>
    group_by(table, step) |>
    summarise(
      n_features     = n(),
      median_cor     = round(median(overlap_cor, na.rm = TRUE), 3),
      pct_cor_gte_09 = round(100 * mean(overlap_cor >= 0.9, na.rm = TRUE), 1),
      pct_cor_lt_08  = round(100 * mean(overlap_cor <  0.8, na.rm = TRUE), 1),
      .groups = "drop"
    )
  knitr::kable(
    summ,
    #caption = "Overlap sample Spearman correlation summary",
    col.names = c("Table", "Step", "Features", "Median cor", "% cor ≥ 0.9", "% cor < 0.8")
  )
}

plot_l2r_hist <- function(df) {
  d <- df[!is.na(df$median_log2ratio), , drop = FALSE]
  if (nrow(d) == 0) { cat("No log2 ratio data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(d, aes(x = median_log2ratio)) +
    geom_histogram(bins = 60, fill = "mediumpurple", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(#title = "Median log2 ratio of overlap sample intensities (batch A / batch B)",
         #subtitle = "Red dashed lines: ±0.5 threshold (≈1.4× fold change)",
         x = "Median log2(batch A / batch B)", y = "Nr. features")
  ggplotly(p)
}

plot_l2r_density <- function(l2r_df) {
  if (is.null(l2r_df) || nrow(l2r_df) == 0) {
    cat("No per-sample log2 ratio data available.\n")
    return(invisible(NULL))
  }
  p <- 
    ggplot(l2r_df, aes(x = log2ratio)) +
    #geom_density(fill = NA, color = "steelblue4", alpha = 0.6, linewidth = 0.4) +
    geom_histogram(bins = 60, fill = "steelblue4", color = "white", linewidth = 0.2) +
    #geom_vline(xintercept = 0, linetype = "solid") +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    #coord_cartesian(xlim = c(-3, 3)) +
    theme_minimal(base_size = 11) +
    labs(#title = "Log2 ratio of overlap sample intensities (batch A / batch B)",
         #subtitle = "Clipped to +-3; ideal distribution is narrow and centred at 0",
         x = "log2(batch A / batch B)", y = "Nr. samples")
  ggplotly(p)
  #print(p)
}

# ── Flagged features ──────────────────────────────────────────────────────────

show_flagged <- function(df) {
  flagged <- df[
    (!is.na(df$overlap_cor)      & df$overlap_cor < 0.8) |
    (!is.na(df$median_log2ratio) & abs(df$median_log2ratio) > 0.5),
    , drop = FALSE
  ]
  if (nrow(flagged) == 0) {
    cat("No features flagged: all matched features pass quality thresholds.\n")
  } else {
    cat(nrow(flagged), "features flagged (low overlap correlation or high intensity ratio).\n\n")
    show_cols <- intersect(
      c("table", "step", "mz", "rt", "identity",
        "mz_ppm", "rt_diff", "overlap_cor", "median_log2ratio"),
      colnames(flagged)
    )
    
    # knitr::kable(
    #   flagged[order(flagged$overlap_cor), show_cols, drop = FALSE],
    #   #caption = "Flagged features — sorted by overlap_cor ascending (up to 100 shown)",
    #   digits  = 4
    # )
    
    datatable(
      flagged[order(flagged$overlap_cor), show_cols, drop = FALSE],
      options = list(
        pageLength = 15,
        autoWidth = TRUE,
        dom = "Bfrtip"
      ),
      filter = "top",
      rownames = FALSE
    )
  }
}


# ── PCA ───────────────────────────────────────────────────────────────────────

plot_pca <- function(tbl, sample_batch_map, title) {
  if (is.null(tbl) || nrow(tbl) == 0) { cat("No data available.\n"); return(invisible(NULL)) }
  int_cols <- if (any(grepl("\\.d\\.area$", colnames(tbl)))) {
    grep("\\.d\\.area$", colnames(tbl), value = TRUE)
  } else {
    grep("(?<!\\.rt|\\.mz)\\.d$", colnames(tbl), value = TRUE, perl = TRUE)
  }
  if (length(int_cols) < 3) { cat("Insufficient sample columns for PCA.\n"); return(invisible(NULL)) }
  mat <- t(as.matrix(tbl[, int_cols, drop = FALSE]))
  mat[is.na(mat)] <- 0
  mat <- log2(mat + 1)
  col_var <- apply(mat, 2, var, na.rm = TRUE)
  mat <- mat[, col_var > 0, drop = FALSE]
  if (ncol(mat) < 2) { cat("Insufficient variable features for PCA.\n"); return(invisible(NULL)) }
  pca  <- prcomp(mat, scale. = TRUE, center = TRUE)
  pct  <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  snames <- sub("^datafile\\.", "", rownames(pca$x))
  snames <- sub("\\.d\\.area$|\\.d$|\\.mzML$", "", snames)
  batch_vals <- sample_batch_map[snames]
  batch_vals[is.na(batch_vals)] <- 0L
  pca_df <- data.frame(
    PC1    = pca$x[, 1],
    PC2    = pca$x[, 2],
    sample = snames,
    batch  = factor(paste0("Batch ", batch_vals),
                    levels = paste0("Batch ", sort(unique(batch_vals)))),
    stringsAsFactors = FALSE
  )
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = batch, label = sample)) +
    geom_point(size = 2.5, alpha = 0.85) +
    theme_minimal(base_size = 11) +
    labs(title = title,
         x = paste0("PC1 (", pct[1], "%)"),
         y = paste0("PC2 (", pct[2], "%)"),
         color = "Batch")
  ggplotly(p, tooltip = c("label", "color", "x", "y"))
}

```

## Inputs

```{r inputs}

cfg         <- wf_read_yaml_config(params$config, label = "params$config")
path_result <- wf_cfg_required(cfg, "path_result")

overlap_names <- if (!wf_is_blank(params$overlap_by_name)) {
  trimws(unlist(strsplit(as.character(params$overlap_by_name), ",", fixed = TRUE)))
} else {
  character(0)
}

diag_path <- file.path(path_result, "data", "merge_diagnostics.rds")
if (!file.exists(diag_path))
  stop("Merge diagnostics not found: ", diag_path,
       "\nRun merge_batches before evaluate_merge.")

diag             <- readRDS(diag_path)
all_diag         <- diag$all_diag
all_counts       <- diag$all_counts
all_l2r          <- diag$all_l2r
all_input_counts <- diag$all_input_counts
n_final_per_tbl  <- diag$n_final

if (is.null(all_diag) || nrow(all_diag) == 0)
  stop("No alignment diagnostics found in ", diag_path)

# Merged peaktables for PCA
tbl_paths <- list(
  pos      = cfg$path_peaktable_pos,
  neg      = cfg$path_peaktable_neg,
  istd_pos = cfg$path_peaktable_istd_pos,
  istd_neg = cfg$path_peaktable_istd_neg
)
merged_tables <- lapply(tbl_paths, function(p) {
  if (wf_is_blank(p) || !file.exists(p)) return(NULL)
  openxlsx::read.xlsx(p)
})

# Sample → batch map: saved by merge_batches into the diagnostics RDS.
sample_batch_map <- if (!is.null(diag$sample_batch_map) && length(diag$sample_batch_map) > 0) {
  diag$sample_batch_map
} else {
  warning("sample_batch_map not found in diagnostics RDS — re-run merge_batches to enable batch colouring")
  integer(0)
}

# Feature subsets
feat_subsets <- make_feature_subsets(all_diag)
l2r_subsets  <- make_l2r_subsets(all_l2r)

cat("Config:                   ", params$config, "\n")
cat("Overlap samples specified:", length(overlap_names), "\n")
cat("Table types:              ", paste(unique(all_diag$table), collapse = ", "), "\n")
cat("Samples mapped to batches:", length(sample_batch_map), "\n")

```

## Alignment diagnostics

### Feature counts

```{r feature_counts, fig.height=4}

merged_counts <- data.frame(
  table    = names(n_final_per_tbl),
  label    = "merged",
  features = unname(n_final_per_tbl),
  stringsAsFactors = FALSE
)

count_plot_df <- rbind(all_input_counts, merged_counts)
input_levels  <- sort(unique(all_input_counts$label))
count_plot_df$label <- factor(count_plot_df$label, levels = c(input_levels, "merged"))

p_counts <- ggplot(count_plot_df, aes(x = label, y = features, fill = ifelse(label == "merged", "After merge", "Before merge"),
                                      text = paste0(label, ": ", features))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("Before merge" = "steelblue", "After merge" = "seagreen")) +
  facet_wrap(~table, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(#title = "Feature counts: input batches vs merged output",
       x = NULL, y = "Nr. features")
ggplotly(p_counts, tooltip = "text")

```

#### Per-step feature counts
```{r feature_counts2, fig.height=4}

knitr::kable(
  all_counts,
  #caption = "Per-step feature counts",
  col.names = c("Table", "Step", "Before merge", "Batch features", "After merge")
)

```

### Alignment quality

#### m/z difference for matched feature pairs

::: {.panel-tabset}

##### Internal standard

```{r align_istd, fig.height=4}
plot_mz_hist(feat_subsets$`Internal standard`)
```

##### With identity

```{r align_id, fig.height=4}
plot_mz_hist(feat_subsets$`With identity`)
```

##### Without identity

```{r align_noid, fig.height=4}
plot_mz_hist(feat_subsets$`Without identity`)
```

##### Overall

```{r align_all, fig.height=4}
plot_mz_hist(feat_subsets$`Overall`)
```

:::

#### Retention time difference for matched feature pairs

::: {.panel-tabset}

##### Internal standard

```{r align_istd2, fig.height=4}
plot_rt_hist(feat_subsets$`Internal standard`)
```

##### With identity

```{r align_id2, fig.height=4}
plot_rt_hist(feat_subsets$`With identity`)
```

##### Without identity

```{r align_noid2, fig.height=4}
plot_rt_hist(feat_subsets$`Without identity`)
```

##### Overall

```{r align_all2, fig.height=4}
plot_rt_hist(feat_subsets$`Overall`)
```

:::

## Overlap sample consistency

### Correlation per feature
::: {.panel-tabset}

#### Internal standard

```{r overlap_istd, fig.height=4}
plot_cor_hist(feat_subsets$`Internal standard`)
print_cor_summary(feat_subsets$`Internal standard`)
```

#### With identity

```{r overlap_id, fig.height=4}
plot_cor_hist(feat_subsets$`With identity`)
print_cor_summary(feat_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid, fig.height=4}
plot_cor_hist(feat_subsets$`Without identity`)
print_cor_summary(feat_subsets$`Without identity`)
```

#### Overall

```{r overlap_all, fig.height=4}
plot_cor_hist(feat_subsets$`Overall`)
print_cor_summary(feat_subsets$`Overall`)
```

:::

### Median ratio of overlap sample intensities
::: {.panel-tabset}

#### Internal standard

```{r overlap_istd2, fig.height=4}

plot_l2r_hist(feat_subsets$`Internal standard`)
```

#### With identity

```{r overlap_id2, fig.height=4}
plot_l2r_hist(feat_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid2, fig.height=4}
plot_l2r_hist(feat_subsets$`Without identity`)
```

#### Overall

```{r overlap_all2, fig.height=4}
plot_l2r_hist(feat_subsets$`Overall`)
```

:::

### Ratio of overlap sample intensities
::: {.panel-tabset}

#### Internal standard

```{r overlap_istd3, fig.height=4}

plot_l2r_density(l2r_subsets$`Internal standard`)
```

#### With identity

```{r overlap_id3, fig.height=4}
plot_l2r_density(l2r_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid3, fig.height=4}
plot_l2r_density(l2r_subsets$`Without identity`)
```

#### Overall

```{r overlap_all3, fig.height=4}
plot_l2r_density(l2r_subsets$`Overall`)
```
:::

## PCA

::: {.panel-tabset}

### Internal standard

```{r pca_is, fig.height=5}
plot_pca(merged_tables$istd_pos, sample_batch_map, "PCA — IS positive mode")
```

```{r pca_is2, fig.height=5}
plot_pca(merged_tables$istd_neg, sample_batch_map, "PCA — IS negative mode")
```

### With identity

```{r pca_id, fig.height=5}
pos_id <- if (!is.null(merged_tables$pos) && "identity" %in% colnames(merged_tables$pos))
  merged_tables$pos[!is.na(merged_tables$pos$identity), , drop = FALSE] else merged_tables$pos

plot_pca(pos_id, sample_batch_map, "PCA — positive mode (identified features)")

```

```{r pca_id2, fig.height=5}
neg_id <- if (!is.null(merged_tables$neg) && "identity" %in% colnames(merged_tables$neg))
  merged_tables$neg[!is.na(merged_tables$neg$identity), , drop = FALSE] else merged_tables$neg

plot_pca(neg_id, sample_batch_map, "PCA — negative mode (identified features)")
```

### Without identity

```{r pca_noid, fig.height=5}
pos_noid <- if (!is.null(merged_tables$pos) && "identity" %in% colnames(merged_tables$pos))
  merged_tables$pos[ is.na(merged_tables$pos$identity), , drop = FALSE] else NULL

plot_pca(pos_noid, sample_batch_map, "PCA — positive mode (unknown features)")

```

```{r pca_noid2, fig.height=5}
neg_noid <- if (!is.null(merged_tables$neg) && "identity" %in% colnames(merged_tables$neg))
  merged_tables$neg[ is.na(merged_tables$neg$identity), , drop = FALSE] else NULL

plot_pca(neg_noid, sample_batch_map, "PCA — negative mode (unknown features)")
```

### Overall

```{r pca_all, fig.height=5}
plot_pca(merged_tables$pos, sample_batch_map, "PCA — positive mode (all features)")
```

```{r pca_all2, fig.height=5}
plot_pca(merged_tables$neg, sample_batch_map, "PCA — negative mode (all features)")
```


:::

## Flagged features

::: {.panel-tabset}

### Internal standard

```{r flag_istd}
show_flagged(feat_subsets$`Internal standard`)
```

### With identity

```{r flag_id}
show_flagged(feat_subsets$`With identity`)
```

### Without identity

```{r flag_noid}
show_flagged(feat_subsets$`Without identity`)
```

### Overall

```{r flag_all}
show_flagged(feat_subsets$`Overall`)
```

:::

```{r session_info, include=FALSE, eval=FALSE}

utils::sessionInfo()

```
