---
title: "Merge batch evaluation (post-processing)"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    toc-location: left
    theme: flatly
    polyfill: false
    page-layout: full
    number-sections: true
    number-depth: 2
self-contained: true

params:
  config: ""
---

```{r setup, include=FALSE, warning=FALSE}

library(yaml)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(openxlsx)

helper_candidates <- c(
  file.path(getwd(), "helpers_workflow.R"),
  file.path(getwd(), "inst", "scripts", "helpers_workflow.R"),
  system.file("scripts", "helpers_workflow.R", package = "metpipe")
)
helper_candidates <- helper_candidates[file.exists(helper_candidates)]
if (length(helper_candidates) == 0) stop("Cannot locate helpers_workflow.R")
source(helper_candidates[[1]], local = TRUE)

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, include = TRUE)

```

```{r functions, include=FALSE}

# ── Feature subset helpers ────────────────────────────────────────────────────

make_feature_subsets <- function(diag_df) {
  # "Known" features (feature_type == "Known") take the role of IS in this workflow
  is_known <- !is.na(diag_df$Feature_type) & diag_df$Feature_type == "Known"
  list(
    #`Known features` = diag_df[ is_known, , drop = FALSE],
    `With identity`  = diag_df[is_known, , drop = FALSE],
    `Without identity` = diag_df[!is_known, , drop = FALSE],
    `Overall`          = diag_df
  )
}

make_l2r_subsets <- function(l2r_df) {
  empty <- data.frame(table = character(), step = character(),
                      identity = character(), sample = character(),
                      log2ratio = numeric(), stringsAsFactors = FALSE)
  if (is.null(l2r_df) || nrow(l2r_df) == 0)
    return(list(`Known features` = empty, `With identity` = empty,
                `Without identity` = empty, `Overall` = empty))
  has_id_col <- "identity" %in% colnames(l2r_df)
  has_ft_col <- "Feature_type" %in% colnames(l2r_df)
  is_known   <- if (has_ft_col) !is.na(l2r_df$Feature_type) & l2r_df$Feature_type == "Known"
                else rep(FALSE, nrow(l2r_df))
  list(
    #`Known features`   = l2r_df[ is_known, , drop = FALSE],
    `With identity`    = l2r_df[is_known, , drop = FALSE],
    `Without identity` = l2r_df[!is_known, , drop = FALSE],
    `Overall`          = l2r_df
  )
}

# ── Alignment quality plots ───────────────────────────────────────────────────

plot_mz_hist <- function(df) {
  if (nrow(df) == 0) { cat("No data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(df, aes(x = mz_ppm)) +
    geom_histogram(bins = 60, fill = "steelblue", color = "white", linewidth = 0.2) +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(x = "m/z difference (ppm)", y = "Nr. features")
  ggplotly(p)
}

plot_rt_hist <- function(df) {
  if (nrow(df) == 0) { cat("No data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(df, aes(x = rt_diff)) +
    geom_histogram(bins = 60, fill = "darkorange", color = "white", linewidth = 0.2) +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(x = "|RT difference| (min)", y = "Nr. features")
  ggplotly(p)
}

plot_ccs_hist <- function(df) {
  if (nrow(df) == 0) { cat("No data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(df, aes(x = ccs_percent)) +
    geom_histogram(bins = 60, fill = "mediumpurple", color = "white", linewidth = 0.2) +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(x = "ccs difference (%)", y = "Nr. features")
  ggplotly(p)
}

# ── Overlap consistency plots ─────────────────────────────────────────────────

plot_cor_hist <- function(df) {
  d <- df[!is.na(df$overlap_cor), , drop = FALSE]
  if (nrow(d) == 0) { cat("No overlap correlation data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(d, aes(x = overlap_cor)) +
    geom_histogram(bins = 60, fill = "seagreen", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = 0.8, linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(subtitle = "Red dashed line: 0.8 quality threshold",
         x = "Spearman r", y = "Nr. features")
  ggplotly(p)
}

print_cor_summary <- function(df) {
  d <- df[!is.na(df$overlap_cor), , drop = FALSE]
  if (nrow(d) == 0) return(invisible(NULL))
  summ <- d |>
    group_by(table, step) |>
    summarise(
      n_features     = n(),
      median_cor     = round(median(overlap_cor, na.rm = TRUE), 3),
      pct_cor_gte_09 = round(100 * mean(overlap_cor >= 0.9, na.rm = TRUE), 1),
      pct_cor_lt_08  = round(100 * mean(overlap_cor <  0.8, na.rm = TRUE), 1),
      .groups = "drop"
    )
  knitr::kable(
    summ,
    col.names = c("Table", "Step", "Features", "Median cor", "% cor ≥ 0.9", "% cor < 0.8")
  )
}

plot_l2r_hist <- function(df) {
  d <- df[!is.na(df$median_log2ratio), , drop = FALSE]
  if (nrow(d) == 0) { cat("No log2 ratio data for this subset.\n"); return(invisible(NULL)) }
  p <- ggplot(d, aes(x = median_log2ratio)) +
    geom_histogram(bins = 60, fill = "mediumpurple", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(x = "Median log2(batch A / batch B)", y = "Nr. features")
  ggplotly(p)
}

plot_l2r_density <- function(l2r_df) {
  if (is.null(l2r_df) || nrow(l2r_df) == 0) {
    cat("No per-sample log2 ratio data available.\n")
    return(invisible(NULL))
  }
  p <- ggplot(l2r_df, aes(x = log2ratio)) +
    geom_histogram(bins = 60, fill = "steelblue4", color = "white", linewidth = 0.2) +
    geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed", color = "firebrick") +
    facet_grid(table ~ step, scales = "free_y") +
    theme_minimal(base_size = 11) +
    labs(x = "log2(batch A / batch B)", y = "Nr. samples")
  ggplotly(p)
}

# ── Flagged features ──────────────────────────────────────────────────────────

show_flagged <- function(df) {
  flagged <- df[
    (!is.na(df$overlap_cor)      & df$overlap_cor < 0.8) |
    (!is.na(df$median_log2ratio) & abs(df$median_log2ratio) > 0.1),
    , drop = FALSE
  ]
  if (nrow(flagged) == 0) {
    cat("No features flagged: all matched features pass quality thresholds.\n")
  } else {
    cat(nrow(flagged), "features flagged (low overlap correlation or high intensity ratio).\n\n")
    show_cols <- intersect(
      c("table", "step", "mz", "rt", "Identity",
        "mz_ppm", "rt_diff", "overlap_cor", "median_log2ratio"),
      colnames(flagged)
    )
    datatable(
      flagged[order(flagged$overlap_cor), show_cols, drop = FALSE],
      extensions = "Buttons",
      options = list(
        pageLength = 15,
        autoWidth = TRUE,
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      ),
      filter = "top",
      rownames = FALSE
    )
  }
}

# ── PCA ───────────────────────────────────────────────────────────────────────

# PCA for datatable format (samples as rows, features as columns).
# dt:              merged datatable data.frame (sample rows)
# fi:              feature.info data.frame
# polarity:        "pos", "neg", or NULL (use both)
# identity_filter: "known" (Feature_type == "Known"), "with_id" (!is.na(Identity)),
#                  "without_id" (is.na(Identity)), or NULL (all)
# sample_batch_map: named integer vector, sample name -> batch number
plot_pca_datatable <- function(dt, fi, polarity = NULL, identity_filter = NULL,
                                sample_batch_map, title) {
  if (is.null(dt) || nrow(dt) == 0) { cat("No data available.\n"); return(invisible(NULL)) }

  # Select features matching polarity + identity_filter
  fi_sub <- fi
  if (!is.null(polarity)) {
    fi_sub <- fi_sub[!is.na(fi_sub$Polarity) & fi_sub$Polarity == polarity, , drop = FALSE]
  }
  if (!is.null(identity_filter)) {
    has_ft  <- "Feature_type" %in% colnames(fi_sub)
    has_id  <- "Identity"     %in% colnames(fi_sub)
    fi_sub  <- switch(identity_filter,
      known       = if (has_ft) fi_sub[!is.na(fi_sub$Feature_type) & fi_sub$Feature_type == "Known", , drop = FALSE]
                    else fi_sub,
      with_id     = if (has_id) fi_sub[!is.na(fi_sub$Identity) & fi_sub$Identity != "", , drop = FALSE]
                    else fi_sub,
      without_id  = if (has_id) fi_sub[ is.na(fi_sub$Identity) | fi_sub$Identity == "", , drop = FALSE]
                    else fi_sub,
      fi_sub
    )
  }
  if (nrow(fi_sub) == 0) { cat("No features for this subset.\n"); return(invisible(NULL)) }

  feat_cols <- intersect(fi_sub$Metabolite_name, colnames(dt))
  if (length(feat_cols) < 3) { cat("Insufficient feature columns for PCA.\n"); return(invisible(NULL)) }

  mat <- as.matrix(dt[, feat_cols, drop = FALSE])
  mat[is.na(mat)] <- 0
  mat <- log2(mat + 1)
  col_var <- apply(mat, 2, var, na.rm = TRUE)
  mat <- mat[, col_var > 0, drop = FALSE]
  if (ncol(mat) < 2) { cat("Insufficient variable features for PCA.\n"); return(invisible(NULL)) }

  pca <- prcomp(mat, scale. = TRUE, center = TRUE)
  pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

  snames    <- dt$Sample
  batch_raw <- sample_batch_map[snames]
  batch_raw[is.na(batch_raw)] <- 0L
  pca_df <- data.frame(
    PC1    = pca$x[, 1],
    PC2    = pca$x[, 2],
    sample = snames,
    batch  = factor(paste0("Batch ", batch_raw),
                    levels = paste0("Batch ", sort(unique(batch_raw)))),
    stringsAsFactors = FALSE
  )
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = batch, label = sample)) +
    geom_point(size = 2.5, alpha = 0.85) +
    theme_minimal(base_size = 11) +
    labs(title = title,
         x = paste0("PC1 (", pct[1], "%)"),
         y = paste0("PC2 (", pct[2], "%)"),
         color = "Batch")
  ggplotly(p, tooltip = c("label", "color", "x", "y"))
}

```

## Inputs

```{r inputs}
#params <- list(config = "E:/Projects/MP208_FLINC/peaktable/PL_LIPS_PASEF/config_win.yml")
cfg          <- wf_read_yaml_config(params$config, label = "params$config")
path_result  <- wf_cfg_required(cfg, "path_result")
path_datatable <- wf_cfg_required(cfg, "path_datatable")

diag_path <- file.path(path_result, "data", "merge_diagnostics_datatable.rds")
if (!file.exists(diag_path))
  stop("Merge diagnostics not found: ", diag_path,
       "\nRun merge_batches_datatable before evaluate_merge_datatable.")

diag             <- readRDS(diag_path)
all_diag         <- diag$all_diag
all_counts       <- diag$all_counts
all_l2r          <- diag$all_l2r
all_input_counts <- diag$all_input_counts
n_final_per_tbl  <- diag$n_final

if (is.null(all_diag) || nrow(all_diag) == 0)
  stop("No alignment diagnostics found in ", diag_path)

# Sample -> batch map saved by merge_batches_datatable
sample_batch_map <- if (!is.null(diag$sample_batch_map) && length(diag$sample_batch_map) > 0) {
  diag$sample_batch_map
} else {
  warning("sample_batch_map not found in diagnostics RDS — re-run merge_batches_datatable")
  integer(0)
}

# Merged datatable for PCA
dt_merged <- if (!wf_is_blank(path_datatable) && file.exists(path_datatable)) {
  openxlsx::read.xlsx(path_datatable, sheet = "datatable")
} else {
  warning("Merged datatable not found: ", path_datatable)
  NULL
}
fi_merged <- if (!wf_is_blank(path_datatable) && file.exists(path_datatable)) {
  openxlsx::read.xlsx(path_datatable, sheet = "feature.info")
} else {
  NULL
}

feat_subsets <- make_feature_subsets(all_diag)
l2r_subsets  <- make_l2r_subsets(all_l2r)

cat("Config:                   ", params$config, "\n")
cat("Table types:              ", paste(unique(all_diag$table), collapse = ", "), "\n")
cat("Samples mapped to batches:", length(sample_batch_map), "\n")

```

## Alignment diagnostics

### Feature counts

```{r feature_counts, fig.height=4}

merged_counts <- data.frame(
  table    = names(n_final_per_tbl),
  label    = "merged",
  features = unname(n_final_per_tbl),
  stringsAsFactors = FALSE
)

count_plot_df <- rbind(all_input_counts, merged_counts)
input_levels  <- sort(unique(all_input_counts$label))
count_plot_df$label <- factor(count_plot_df$label, levels = c(input_levels, "merged"))

p_counts <- ggplot(count_plot_df, aes(x = label, y = features,
                                      fill = ifelse(label == "merged", "After merge", "Before merge"),
                                      text = paste0(label, ": ", features))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("Before merge" = "steelblue", "After merge" = "seagreen")) +
  facet_wrap(~table, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = NULL, y = "Nr. features")
ggplotly(p_counts, tooltip = "text")

```

#### Per-step feature counts
```{r feature_counts2, fig.height=4}

knitr::kable(
  all_counts,
  col.names = c("Table", "Step", "Before merge", "Batch features", "After merge")
)

```

### Alignment quality

#### m/z difference for matched feature pairs

::: {.panel-tabset}

##### With identity

```{r align_id, fig.height=4}
plot_mz_hist(feat_subsets$`With identity`)
```

##### Without identity

```{r align_noid, fig.height=4}
plot_mz_hist(feat_subsets$`Without identity`)
```

##### Overall

```{r align_all, fig.height=4}
plot_mz_hist(feat_subsets$`Overall`)
```

:::

#### Retention time difference for matched feature pairs

::: {.panel-tabset}

##### With identity

```{r align_id2, fig.height=4}
plot_rt_hist(feat_subsets$`With identity`)
```

##### Without identity

```{r align_noid2, fig.height=4}
plot_rt_hist(feat_subsets$`Without identity`)
```

##### Overall

```{r align_all2, fig.height=4}
plot_rt_hist(feat_subsets$`Overall`)
```

:::

#### ccs difference for matched feature pairs

::: {.panel-tabset}

##### With identity

```{r align_id3, fig.height=4}
plot_ccs_hist(feat_subsets$`With identity`)
```

##### Without identity

```{r align_noid3, fig.height=4}
plot_ccs_hist(feat_subsets$`Without identity`)
```

##### Overall

```{r align_all3, fig.height=4}
plot_ccs_hist(feat_subsets$`Overall`)
```

:::

## Overlap sample consistency

### Correlation per feature
::: {.panel-tabset}

#### With identity

```{r overlap_id, fig.height=4}
plot_cor_hist(feat_subsets$`With identity`)
print_cor_summary(feat_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid, fig.height=4}
plot_cor_hist(feat_subsets$`Without identity`)
print_cor_summary(feat_subsets$`Without identity`)
```

#### Overall

```{r overlap_all, fig.height=4}
plot_cor_hist(feat_subsets$`Overall`)
print_cor_summary(feat_subsets$`Overall`)
```

:::

### Median ratio of overlap sample intensities
::: {.panel-tabset}

#### With identity

```{r overlap_id2, fig.height=4}
plot_l2r_hist(feat_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid2, fig.height=4}
plot_l2r_hist(feat_subsets$`Without identity`)
```

#### Overall

```{r overlap_all2, fig.height=4}
plot_l2r_hist(feat_subsets$`Overall`)
```

:::

### Ratio of overlap sample intensities
::: {.panel-tabset}

#### With identity

```{r overlap_id3, fig.height=4}
plot_l2r_density(l2r_subsets$`With identity`)
```

#### Without identity

```{r overlap_noid3, fig.height=4}
plot_l2r_density(l2r_subsets$`Without identity`)
```

#### Overall

```{r overlap_all3, fig.height=4}
plot_l2r_density(l2r_subsets$`Overall`)
```
:::

## PCA

::: {.panel-tabset}

### With identity

```{r pca_id_pos, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "pos", identity_filter = "with_id",
                   sample_batch_map = sample_batch_map,
                   title = "PCA — positive mode (identified features)")
```

```{r pca_id_neg, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "neg", identity_filter = "with_id",
                   sample_batch_map = sample_batch_map,
                   title = "PCA — negative mode (identified features)")
```

### Without identity

```{r pca_noid_pos, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "pos", identity_filter = "without_id",
                   sample_batch_map = sample_batch_map,
                   title = "PCA — positive mode (unknown features)")
```

```{r pca_noid_neg, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "neg", identity_filter = "without_id",
                   sample_batch_map = sample_batch_map,
                   title = "PCA — negative mode (unknown features)")
```

### Overall

```{r pca_all_pos, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "pos", identity_filter = NULL,
                   sample_batch_map = sample_batch_map,
                   title = "PCA — positive mode (all features)")
```

```{r pca_all_neg, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = "neg", identity_filter = NULL,
                   sample_batch_map = sample_batch_map,
                   title = "PCA — negative mode (all features)")
```

### Combined pos + neg

```{r pca_combined, fig.height=5}
plot_pca_datatable(dt_merged, fi_merged, polarity = NULL, identity_filter = NULL,
                   sample_batch_map = sample_batch_map,
                   title = "PCA — positive + negative mode (all features)")
```

:::

## Flagged features

::: {.panel-tabset}

### With identity

```{r flag_id}
show_flagged(feat_subsets$`With identity`)
```

### Without identity

```{r flag_noid}
show_flagged(feat_subsets$`Without identity`)
```

### Overall

```{r flag_all}
show_flagged(feat_subsets$`Overall`)
```

:::

```{r session_info, include=FALSE, eval=FALSE}

utils::sessionInfo()

```
