---
title: "Remove theoretical annotation"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '3'
    number_sections: true
    df_print: paged

params:
  config: ""
---
<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE, warning = FALSE}

library(dplyr)
library(tibble)
library(plotly)
library(ggsci)
library(yaml)
library(openxlsx)

load_metpipe_runtime <- function() {
  package_mode <- tolower(trimws(Sys.getenv("METPIPE_PACKAGE_MODE", unset = "local")))
  if (!package_mode %in% c("local", "installed")) package_mode <- "local"

  if (identical(package_mode, "installed")) {
    if (!requireNamespace("metpipe", quietly = TRUE)) {
      stop("METPIPE_PACKAGE_MODE='installed' but package 'metpipe' is not installed")
    }
    library(metpipe)
    return("installed")
  }

  candidates <- unique(Filter(nzchar, c(
    Sys.getenv("METPIPE_LOCAL_ROOT", ""),
    "H:/Documents/CBMR_workflow/packages/metpipe",
    normalizePath(getwd(), winslash = "/", mustWork = FALSE),
    normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), winslash = "/", mustWork = FALSE)
  )))

  is_pkg_root <- function(path) {
    dir.exists(path) &&
      file.exists(file.path(path, "DESCRIPTION")) &&
      dir.exists(file.path(path, "R"))
  }

  roots <- candidates[vapply(candidates, is_pkg_root, logical(1))]
  if (length(roots) == 0) {
    stop("Cannot locate local metpipe package root. Set METPIPE_LOCAL_ROOT to your metpipe folder.")
  }

  if (!requireNamespace("pkgload", quietly = TRUE)) {
    stop("pkgload is required for local source loading. Install with install.packages('pkgload').")
  }

  Sys.setenv(METPIPE_LOCAL_ROOT = roots[[1]])
  pkgload::load_all(path = roots[[1]], export_all = FALSE, quiet = TRUE)
  "local"
}

metpipe_runtime <- load_metpipe_runtime()

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

is_blank <- function(x) {
  is.null(x) || length(x) == 0 || all(is.na(x)) || !nzchar(trimws(as.character(x[1])))
}

cfg_required <- function(cfg, name) {
  val <- cfg[[name]]
  if (is_blank(val)) stop("Missing required config field: ", name)
  val
}
```

```{r pipeline input}

if (is_blank(params$config)) {
  stop("params$config is required")
}
if (!file.exists(params$config)) {
  stop("Config file does not exist: ", params$config)
}

config <- read_yaml(params$config)
if (is.null(config) || !is.list(config)) {
  stop("Invalid YAML config: expected a named list")
}

rds_datalist <- cfg_required(config, "rds_datalist")
path_datatable <- cfg_required(config, "path_datatable")
path_workflow = cfg_required(config, "path_workflow")
final.eval.sample.to.use = cfg_required(config, "final.eval.sample.to.use")

if (!file.exists(rds_datalist)) {
  stop("datalist file not found: ", rds_datalist)
}

db_file <- file.path(path_workflow, "library", "metabolomics_workbench_refmet_20250515.csv")
if (!file.exists(db_file)) {
  stop("Metabolite mapping database file not found: ", db_file)
}

out_dir <- dirname(path_datatable)
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

```

```{r, eval = FALSE, include = FALSE}
rds_datalist <- "D:/Projects/MP197/peaktable/_BL_LIPS_FPASEF/datalist.rds"

```

# Calculate sum composition
```{r}

datalist <- readRDS(rds_datalist)

if (!is.list(datalist) || is.null(datalist$datatable) || is.null(datalist$sample.info) || is.null(datalist$feature.info)) {
  stop("Invalid datalist: expected datatable, sample.info, and feature.info")
}

# change theoretical annotation to unknown
sample.info <- datalist$sample.info
feature.info <- datalist$feature.info
datatable <- datalist$datatable
peaks <- datatable[, (ncol(sample.info) + 1):ncol(datatable)]

if (ncol(peaks) == 0) {
  stop("No feature columns found in datalist$datatable")
}


feature.info <- 
  add_lipid_info(feature.info) %>% 
  relocate(c(Identity_sum, Rep), .before = Metabolite_name) %>% 
  rename(Metabolite_name_chain = Metabolite_name) %>% 
  select(-c(refmet_id:inchi_key))

mean_col <- paste0("mean.", final.eval.sample.to.use)
if (!mean_col %in% names(feature.info)) {
  stop("Required evaluation column not found in feature.info: ", mean_col)
}

if (!"Metabolite_name_chain" %in% names(feature.info)) {
  stop("Expected column missing after lipid parsing: Metabolite_name_chain")
}

```

# Remove repeating lipids with the same sum composition
Lipids with the same sum composition and similar retention time are considered 
potentially the same lipids. Only one lipid per such group is kept.
```{r}
feature.info.clean <- 
  feature.info %>% 
  group_by(Identity_sum) %>%
  arrange(Identity_sum, desc(get(mean_col))) %>%
  mutate(first_rt = first(rt),
         diff_to_first_rt = abs(rt - first_rt) > 0.2) %>% 
  filter(row_number() == 1 | diff_to_first_rt == TRUE) %>% 
  ungroup() %>% 
  select(-c(first_rt, diff_to_first_rt))

keep_cols <- intersect(feature.info.clean$Metabolite_name_chain, colnames(peaks))
if (length(keep_cols) == 0) {
  stop("No matching feature columns found for cleaned lipid identities")
}
peaks.clean <- peaks[, keep_cols, drop = FALSE]

cat("For repeating features, the one with higher mean intensity in ", final.eval.sample.to.use, " was kept: \n")
cat(nrow(feature.info) - nrow(feature.info.clean), " features have been removed")
```

# Map lipid information
```{r}

feature.info.clean.map <- 
  map_metabolite_info(input_file = feature.info.clean,
                      name_col = match("Identity_sum", names(feature.info.clean)),
                      sep = ';',
                      db_file = db_file) %>% 
  select(-Metabolite_name_original)

```

# Remove theoretical annotation
```{r}

feature.info.no.theo <- 
  feature.info.clean.map %>% 
  mutate(flag = grepl("*", Metabolite_name, fixed = TRUE),
         Feature_type = ifelse(flag, "Unknown", Feature_type),
         Identity = ifelse(flag, paste0("Unknown_MZ_", round(mz, 3), "_RT_", round(rt, 3)), Identity),
         identity_source = ifelse(flag, NA, identity_source),
         Metabolite_name = ifelse(flag, Identity, Metabolite_name)) %>% 
  group_by(Metabolite_name) %>% 
  mutate(n = n(),
         Metabolite_name = if_else(n > 1, paste0(Metabolite_name, "-iso", row_number()), Metabolite_name)) %>% 
  ungroup() %>% 
  arrange(identity_source, Metabolite_name) %>% 
  select(-c(flag, n, Rep)) %>% 
  relocate(Metabolite_name, .after = mz)

keep_cols_new <- intersect(feature.info.no.theo$Metabolite_name_chain, colnames(peaks.clean))
peaks.new <- peaks.clean[, keep_cols_new, drop = FALSE]

feature.info.no.theo <- feature.info.no.theo[match(keep_cols_new, feature.info.no.theo$Metabolite_name_chain), , drop = FALSE]

names(peaks.new) <- feature.info.no.theo$Metabolite_name

datatable.new <- cbind(sample.info, peaks.new)

datalist$datatable.no.theo <- datatable.new
datalist$feature.info.no.theo <- feature.info.no.theo

saveRDS(datalist, rds_datalist)
```


# Export
```{r}
wb <- createWorkbook()
addWorksheet(wb, "datatable")
writeData(wb, "datatable", datalist$datatable.no.theo)
addWorksheet(wb, "feature.info")
writeData(wb, "feature.info", datalist$feature.info.no.theo)
saveWorkbook(wb, path_datatable, overwrite = TRUE)

```


```{r session info, include = FALSE}

# Session info

utils::sessionInfo()

```