---
title: "Remove theoretical annotation"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '3'
    number_sections: true
    df_print: paged

params:
  config: ""
---
<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE, warning = FALSE}

library(dplyr)
library(tibble)
library(plotly)
library(metpipe)
library(ggsci)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)
```

```{r pipeline input}

config <- read_yaml(params$config)

rds_datalist <- config$rds_datalist
path_datatable <- config$path_datatable
path_workflow = config$path_workflow
final.eval.sample.to.use = config$final.eval.sample.to.use

```

```{r, eval = FALSE, include = FALSE}
rds_datalist <- "D:/Projects/MP197/peaktable/_BL_LIPS_FPASEF/datalist.rds"

```

# Calculate sum composition
```{r}

datalist <- readRDS(rds_datalist)

# change theoretical annotation to unknown
sample.info <- datalist$sample.info
feature.info <- datalist$feature.info
datatable <- datalist$datatable
peaks <- datatable[, (ncol(sample.info) + 1):ncol(datatable)]


feature.info <- 
  add_lipid_info(feature.info) %>% 
  relocate(c(Identity_sum, Rep), .before = Metabolite_name) %>% 
  rename(Metabolite_name_chain = Metabolite_name) %>% 
  select(-c(refmet_id:inchi_key))

```

# Remove repeating lipids with the same sum composition
Lipids with the same sum composition and similar retention time are considered 
potentially the same lipids. Only one lipid per such group is kept.
```{r}
feature.info.clean <- 
  feature.info %>% 
  group_by(Identity_sum) %>%
  arrange(Identity_sum, desc(get(paste0("mean.", final.eval.sample.to.use)))) %>%
  mutate(first_rt = first(rt),
         diff_to_first_rt = abs(rt - first_rt) > 0.2) %>% 
  filter(row_number() == 1 | diff_to_first_rt == TRUE) %>% 
  ungroup() %>% 
  select(-c(first_rt, diff_to_first_rt))

peaks.clean <- peaks[, feature.info.clean$Metabolite_name_chain]

cat("For repeating features, the one with higher mean intensity in ", eval.sample.to.use, " was kept: \n")
cat(nrow(feature.info) - nrow(feature.info.clean), " features have been removed")
```

# Map lipid information
```{r}

feature.info.clean.map <- 
  map_metabolite_info(input_file = feature.info.clean,
                      name_col = 3,
                      sep = ';',
                      db_file = paste0(path_workflow, '/library/metabolomics_workbench_refmet_20250515.csv')) %>% 
  select(-Metabolite_name_original)

```

# Remove theoretical annotation
```{r}

feature.info.no.theo <- 
  feature.info.clean.map %>% 
  mutate(flag = grepl("*", Metabolite_name, fixed = TRUE),
         Feature_type = ifelse(flag, "Unknown", Feature_type),
         Identity = ifelse(flag, paste0("Unknown_MZ_", round(mz, 3), "_RT_", round(rt, 3)), Identity),
         identity_source = ifelse(flag, NA, identity_source),
         Metabolite_name = ifelse(flag, Identity, Metabolite_name)) %>% 
  group_by(Metabolite_name) %>% 
  mutate(n = n(),
         Metabolite_name = if_else(n > 1, paste0(Metabolite_name, "-iso", row_number()), Metabolite_name)) %>% 
  ungroup() %>% 
  arrange(identity_source, Metabolite_name) %>% 
  select(-c(flag, n, Rep))

peaks.new <- peaks.clean[, feature.info.no.theo$Metabolite_name_chain]

names(peaks.new) <- feature.info.no.theo$Metabolite_name

datatable.new <- cbind(sample.info, peaks.new)

datalist$datatable.no.theo <- datatable.new
datalist$feature.info.no.theo <- feature.info.no.theo

saveRDS(datalist, rds_datalist)
```


# Export
```{r}
wb <- createWorkbook()
addWorksheet(wb, "datatable")
writeData(wb, "datatable", datalist$datatable.no.theo)
addWorksheet(wb, "feature.info")
writeData(wb, "feature.info", datalist$feature.info.no.theo)
saveWorkbook(wb, path_datatable, overwrite = TRUE)

```


```{r session info, include = FALSE}

# Session info

utils::sessionInfo()

```