---
title: "Prepare parameters for workflow"
author: "CBMR Metabolomics Platform"
date: "`r base::Sys.Date()`"
          
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: '3'
    number_sections: true
    df_print: paged

params:
  path_raw: ""
  path_result: ""
  path_workflow: ""
  path_mzmine: ""
  path_temp: ""
---
<style type="text/css">
body{
  font-size: 12px;
  }
td {
  font-size: 14px;
}
h1.title {
  font-size: 22px;
}
h4.author {
  font-size: 16px;
}
h4.date {
  font-size: 16px;
}

h1 {
  font-size: 18px;
}
h2 {
  font-size: 16px;
}
h3 {
  font-size: 14px;
}
h4 {
  font-size: 12px;
}
h5 {
  font-size: 12px;
}
code.r{
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

```{r setup environment, include = FALSE}

library(yaml)
library(dplyr)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  include = TRUE
)

```

# Global input
```{r}

# path_raw = "F:/Projects/MP207/raw/ME_LIPL_FDDA"
# path_result = "F:/Projects/MP207/peaktable/ME_LIPL_FDDA"
path_raw = "F:/Projects/MP207/raw/ME_zHILIC_FDDA"
path_result = "F:/Projects/MP207/peaktable/ME_zHILIC_FDDA"
path_workflow = "E:/Qian/workflow"
path_mzmine = "E:/Qian/workflow/mzmine_windows/mzmine_console.exe"
path_temp = "F:/Temp"

if (!dir.exists(path_result)) dir.create(path_result)

```

```{r, eval = FALSE}

path_raw = params$path_raw
path_result = params$path_result
path_workflow = params$path_workflow
path_mzmine = params$path_mzmine
path_temp = params$path_temp

```

# Workflow selection
```{r}

method_all <- readRDS(paste0(path_workflow, "/parameters/method_configuration.rds"))

method_name <- basename(path_raw)
matrix_i <- strsplit(method_name, "_")[[1]][1]
assay_i <- strsplit(method_name, "_")[[1]][2]
analysis_i <- strsplit(method_name, "_")[[1]][3]

method <- subset(method_all, assay == assay_i & analysis == analysis_i)

```

# Raw data files
```{r}
path_files_pos <- paste0(path_raw, "/pos/")
path_files_neg <- paste0(path_raw, "/neg/")

run.pos <- FALSE
run.neg <- FALSE
raw_files_pos <- NULL
raw_files_neg <- NULL
meta_pos <- NULL
meta_neg <- NULL

if (dir.exists(path_files_pos)){
  run.pos <- TRUE
  raw_files_pos <- list.files(path_files_pos, full.names = TRUE, pattern = ".d$")
  meta_pos <- map_standard_name_to_meta(raw_files_pos)
    #metpipe::map_standard_name_to_meta(raw_files_pos) 
}

if (dir.exists(path_files_neg)){
  run.neg <- TRUE
  raw_files_neg <- list.files(path_files_neg, full.names = TRUE, pattern = ".d$")
  meta_neg <- map_standard_name_to_meta(raw_files_neg)
    #metpipe::map_standard_name_to_meta(raw_files_neg) 
}


```

# mzmine input
```{r}

#path_result <- gsub("raw", "peaktable", path_raw)
mzmine_params <- paste0(path_workflow, method$mzmine_params)
path_lib_istd_pos <- ifelse(!is.na(method$path_lib_istd_pos), paste0(path_workflow, method$path_lib_istd_pos), NA)
path_lib_istd_neg <- ifelse(!is.na(method$path_lib_istd_neg), paste0(path_workflow, method$path_lib_istd_neg), NA)
# path_lib_pos <- ifelse(!is.na(method$path_lib_pos), paste0(path_workflow, method$path_lib_pos), NA)
# path_lib_neg <- ifelse(!is.na(method$path_lib_neg), paste0(path_workflow, method$path_lib_neg), NA)
spectral_lib <- strsplit(method$path_spectral, "|", fixed = TRUE)[[1]]
path_spectral_lib <- paste0(path_workflow, spectral_lib)
mzmine_user <- paste0(path_workflow, "/parameters/qiangao.mzuser")
mzmine_memory <- "-memory none"
mzmine_threads <- "-threads auto"
```

# QC
```{r}
figure_height = 800
figure_width = 1000

```

# post processing
## Clean peaktable
```{r}

mean.filter <- list(c( 2, "BL", "<", "PO"),
                   c( 1, "Sample", ">", 500))
rsd.filter <- list(c(1, "PO", "<", 0.3))
rt.range <- c(0, 30)
filter_by_missing_feature_pct <- 50
filter_by_missing_sample_type <- "Sample"
outliers.sample <- c()
clean.po.sample.to.use <- "PO"

```

## Normalization
```{r}

impute.method.sample <- "HF"
impute.method.is <- "HF"
norm.po.sample.to.use <- "PO"
norm.method <- c("bestis", "low_cv", "sum", "median")
sample.type.keep <- c("NIST", "PO")

```

## Merge and map info
```{r}

final.norm <- "bestis"
final.eval.sample.to.use <- "PO"
final.sample.type.keep <- c("Sample", "PO", "NIST")
```

## Others
```{r}

mzmine_peaktable_pos <- paste0(path_result, "/pos/pos_full_feature_table.csv")
mzmine_peaktable_neg <- paste0(path_result, "/neg/neg_full_feature_table.csv")
path_peaktable_pos <- paste0(path_result, "/peaktable_pos.xlsx")
path_peaktable_neg <- paste0(path_result, "/peaktable_neg.xlsx")
path_peaktable_istd_pos <- paste0(path_result, "/peaktable_istd_pos.xlsx")
path_peaktable_istd_neg <- paste0(path_result, "/peaktable_istd_neg.xlsx")
if (!dir.exists(paste0(path_result, "/data"))) dir.create(paste0(path_result, "/data"))
rds_qc_is_pos <- paste0(path_result, "/data/QC_IS_pos.rds")
rds_qc_is_neg <- paste0(path_result, "/data/QC_IS_neg.rds")
rds_datalist <- paste0(path_result, "/datalist.rds")
path_datatable <- paste0(path_result, "/datatable.xlsx")
```

# Generate yml
```{r}

config <- list(
  path_raw = path_raw,
  path_workflow = path_workflow,
  path_mzmine = path_mzmine,
  mzmine_threads = mzmine_threads,
  path_temp = path_temp,
  method_name = method_name,
  run.pos = run.pos,
  run.neg = run.neg,  
  path_result = path_result, 
  mzmine_params = mzmine_params, 
  path_lib_istd_pos = path_lib_istd_pos, 
  path_lib_istd_neg = path_lib_istd_neg,  
  path_spectral_lib = path_spectral_lib,  
  figure_height = figure_height,
  figure_width = figure_width,  
  mean.filter = mean.filter,
  rsd.filter = rsd.filter,
  rt.range = rt.range,
  filter_by_missing_feature_pct = filter_by_missing_feature_pct,
  filter_by_missing_sample_type = filter_by_missing_sample_type,
  outliers.sample = outliers.sample,
  clean.po.sample.to.use = clean.po.sample.to.use,
  impute.method.sample = impute.method.sample,
  impute.method.is = impute.method.is,
  norm.po.sample.to.use = norm.po.sample.to.use,
  norm.method = norm.method,
  sample.type.keep = sample.type.keep,
  final.norm = final.norm,
  final.eval.sample.to.use = final.eval.sample.to.use,
  final.sample.type.keep = final.sample.type.keep,  
  
  raw_files_pos = raw_files_pos,
  meta_pos = meta_pos,
  raw_files_neg = raw_files_neg,
  meta_neg = meta_neg,
  # path_lib_pos = path_lib_pos,
  # path_lib_neg = path_lib_neg,
  spectral_lib = spectral_lib,
  mzmine_user = mzmine_user,
  mzmine_memory = mzmine_memory,
  mzmine_peaktable_pos = mzmine_peaktable_pos,
  mzmine_peaktable_neg = mzmine_peaktable_neg,
  path_peaktable_pos = path_peaktable_pos,
  path_peaktable_neg = path_peaktable_neg,
  path_peaktable_istd_pos = path_peaktable_istd_pos,
  path_peaktable_istd_neg = path_peaktable_istd_neg,
  rds_qc_is_pos = rds_qc_is_pos,
  rds_qc_is_neg = rds_qc_is_neg,
  rds_datalist = rds_datalist,
  path_datatable = path_datatable
)

yaml_str <- as.yaml(config)
cat(yaml_str)
writeLines(yaml_str, paste0(path_result, "/config.yml"))
```

```{r}

```













